---
title: "Analysis Notebook"
toc: true
toc-depth: 3
number-sections: true
fig-width: 8
fig-height: 6

execute:
  freeze: false
  cache: false

---

# Setup
```{r}
#| label: setup
#| include: false

packages <- c("DT","MuMIn","lunar","glmmTMB","DHARMa","broom.mixed","emmeans","lmerTest","lme4","patchwork","tidyverse","writexl","readxl")

quiet_load <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    suppressWarnings(suppressMessages(install.packages(pkg, dependencies = c("Depends", "Imports", "LinkingTo"))))
  }
  suppressPackageStartupMessages(require(pkg, character.only = TRUE, quietly = TRUE))
  invisible(TRUE)
}

options(repos = c(CRAN = "https://cloud.r-project.org"))
invisible(lapply(packages, quiet_load))

exc_file_dir  <- "data/raw/Data_Stone_piles_Rotoiti.xlsx"
raw_data_dir  <- "data/raw"
der_data_dir  <- "data/derived"
out_dir       <- "outputs"

# Read data
Site_info          <- readxl::read_excel(exc_file_dir, sheet = "Site_info")
Monitoring_data    <- readxl::read_excel(exc_file_dir, sheet = "Monitoring_data")
Chemical_data      <- readxl::read_excel(exc_file_dir, sheet = "Chemical_data")
Stone_pile_data    <- readxl::read_excel(exc_file_dir, sheet = "Stone_pile_data")  %>% select(-starts_with("..."))
Weed_data          <- readxl::read_excel(exc_file_dir, sheet = "Weed_data")  %>% select(-starts_with("..."))
Fish_data          <- readxl::read_excel(exc_file_dir, sheet = "Fish_data")  %>% select(-starts_with("..."))
Macroinvertebrates <- readxl::read_excel(exc_file_dir, sheet = "Macroinvertebrates")

# Save as csv to derived
write.csv(Site_info,          file.path(der_data_dir, "Site_info.csv"),          row.names = FALSE)
write.csv(Monitoring_data,    file.path(der_data_dir, "Monitoring_data.csv"),    row.names = FALSE)
write.csv(Chemical_data,      file.path(der_data_dir, "Chemical_data.csv"),      row.names = FALSE)
write.csv(Stone_pile_data,    file.path(der_data_dir, "Stone_pile_data.csv"),    row.names = FALSE)
write.csv(Weed_data,          file.path(der_data_dir, "Weed_data.csv"),          row.names = FALSE)
write.csv(Fish_data,          file.path(der_data_dir, "Fish_data.csv"),          row.names = FALSE)
write.csv(Macroinvertebrates, file.path(der_data_dir, "Macroinvertebrates.csv"), row.names = FALSE)



Site_info          <- read.csv(file.path(der_data_dir, "Site_info.csv"))
Monitoring_data    <- read.csv(file.path(der_data_dir, "Monitoring_data.csv"))
Chemical_data      <- read.csv(file.path(der_data_dir, "Chemical_data.csv"))
Stone_pile_data    <- read.csv(file.path(der_data_dir, "Stone_pile_data.csv"))
Weed_data          <- read.csv(file.path(der_data_dir, "Weed_data.csv"))
Fish_data          <- read.csv(file.path(der_data_dir, "Fish_data.csv"))
Macroinvertebrates <- read.csv(file.path(der_data_dir, "Macroinvertebrates.csv"))

site_colours <- c("Stone_pile" = "#D55E00", "Control" = "#E69F00", "Reference" = "#0072B2")
site_colours_effect <- c("Stone pile" = "#D55E00", "Control" = "#E69F00", "Reference1" = "#0072B2", "Reference2" = "#56B4E9")
site_fill <- c("Stone_pile" = "#D55E00", "Control" = "#E69F00", "Reference" = "#0072B2")
site_shapes  <- c("Stone_pile" = 21,  "Control" = 24, "Reference" = 22)
site_shapes_effect <- c("Stone pile" = 21,  "Control" = 24, "Reference1" = 22, "Reference2" = 22)

conv_colours <- c("Stone_pile"= "#D55E00","Reference" = "#0072B2")
conv_fill    <- conv_colours
conv_shapes  <- c("Stone_pile" = 21,"Reference" = 22)

site_labels <- c("Stone_pile" = "Stone pile", "Control" = "Control", "Reference" = "Reference")
conv_labels <- c("Stone_pile" = "Stone pile", "Reference" = "Reference")

season_colours <- c(Summer = "#D55E00", Autumn = "#E69F00", Winter = "#56B4E9", Spring = "#009E73")
season_shapes  <- c(Summer = 21, Autumn = 24, Winter = 22, Spring = 23)

monitor_labels <- c("Summer\n2025", "Autumn\n2025", "Winter\n2025", "Spring\n2025", "Summer\n2026", "Autumn\n2026", "Winter\n2026")

base_theme_bw <- theme_classic() +
  theme(text     = element_text(family = "Arial", size = 11),
    axis.title   = element_text(face = "plain"),
    axis.text    = element_text(face = "plain"),
    plot.title   = element_text(face = "plain"),
    strip.text   = element_text(face = "plain"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.3))
theme_set(base_theme_bw)

format_pval <- function(p, label = TRUE) {
  out <- if (p < 0.001) "< 0.001" else paste0("= ", format(round(p, 3), nsmall = 3))
  if (label) paste0("*p* ", out) else out
}


stat <- function(mid, trm, col) {
  tidy_all_full |>
    dplyr::filter(model_id == mid,
                  stringr::str_detect(term, trm)) |>
    dplyr::pull({{ col }}) |>
    dplyr::first()
}

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 0.001) return("*p* < 0.001")
  paste0("*p* = ", formatC(p, format = "f", digits = 3))
}

fmt_b <- function(x, digits = 2) {
  formatC(x, format = "f", digits = digits)
}

fmt_num <- function(x, digits = 3) {
  formatC(x, format = "f", digits = digits)
}

fmt_ci <- function(mid, trm, digits = 2) {
  row <- tidy_all_full |>
    dplyr::filter(model_id == mid, stringr::str_detect(term, trm))
  lo <- formatC(dplyr::pull(row, conf.low),  format = "f", digits = digits)
  hi <- formatC(dplyr::pull(row, conf.high), format = "f", digits = digits)
  paste0("[", lo, ", ", hi, "]")
}

stat_ms <- function(sp, rt, col) {
  matched_season_all |>
    dplyr::filter(species == sp, response_type == rt) |>
    dplyr::pull({{ col }}) |>
    dplyr::first()
}


```

# Data ingestion and preparation
## Length–weight
```{r}
#| label: fig-length-weight
#| include: true
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Length–weight relationship for kōura, fitted on log₁₀–log₁₀ scale. Open points are weights predicted from the length–weight regression (used to fill missing scale measurements); filled points are directly measured weights."

lw_model <- lm(log10(weight_g) ~ log10(length_mm), data = Fish_data, subset = species_name == "Freshwater_crayfish", na.action = na.exclude)

sigma_log10 <- sigma(lw_model)
c <- 10^(0.5 * sigma_log10^2)

Fish_data <- Fish_data |>
  mutate(is_koura = species_name == "Freshwater_crayfish",
    Predicted_weight = case_when( !is.na(weight_g) ~ weight_g,
      is_koura ~ {pred_log <- predict(lw_model, newdata = pick(everything()))
        10^(pred_log) * c}, TRUE ~ NA_real_ ),
    Weight_source = case_when(is_koura & !is.na(weight_g) ~ "measured",
      is_koura & is.na(weight_g) ~ "predicted",TRUE ~ NA_character_))

count_df <- Fish_data |>
  dplyr::filter(is_koura, !is.na(Weight_source)) |>
  dplyr::count(Weight_source, name = "n")

measured_n  <- count_df$n[count_df$Weight_source == "measured"]
predicted_n <- count_df$n[count_df$Weight_source == "predicted"]

total_koura <- Fish_data |>
  dplyr::filter(is_koura) |>
  nrow()

hist(Fish_data |> dplyr::filter(is_koura) |> dplyr::pull(length_mm), xlab = "Length (mm)")

a <- coef(lw_model)[1]
b <- coef(lw_model)[2]
r2_lw <- summary(lw_model)$r.squared
formula_text <- paste0( "log10(Weight[g]) = ", round(a, 3), " + ", round(b, 3), " * log10(OCL Length[mm])",
  "\nR² = ", round(r2_lw, 3), ", n = ", measured_n)

length_weight_plot <- ggplot(
Fish_data |> dplyr::filter(is_koura),
aes(x = length_mm, y = Predicted_weight, shape = Weight_source)) +
geom_point(size = 2.2, colour = "black", stroke = 0.6) +
scale_shape_manual(values = c("measured" = 16, "predicted" = 1)) +
labs(x = "OCL length (mm)", y = "Weight (g)", shape = "Weight source") +
annotate("text", x = Inf, y = Inf, label = formula_text, hjust = 1.1, vjust = 1.3, size = 3) 

ggsave(file.path(out_dir, "fig-length-weight-ch5.png"),length_weight_plot,width = 8, height = 5, dpi = 300)

length_weight_plot

```

## Build CPUE and BPUE
```{r}
#| label: cpue-bpue
#| include: false

CPUE_BPUE_legacy <- Fish_data %>%
  filter(!is.na(species)) %>%
  group_by(monitoring_id, species, net_type) %>%
  dplyr::reframe(
    Total_Individuals = sum(amount, na.rm = TRUE),
    Total_Weight      = sum(weight_g, na.rm = TRUE),
    Total_Effort      = dplyr::first(amount_nets),
    CPUE              = Total_Individuals / Total_Effort,
    BPUE              = Total_Weight      / Total_Effort,
    Mean_Length       = mean(length_mm, na.rm = TRUE),
    Min_Length        = ifelse(all(is.na(length_mm)), NA, min(length_mm, na.rm = TRUE)),
    Max_Length        = ifelse(all(is.na(length_mm)), NA, max(length_mm, na.rm = TRUE)),
    Mean_Weight       = mean(weight_g, na.rm = TRUE),
    Min_Weight        = ifelse(all(is.na(weight_g)), NA, min(weight_g, na.rm = TRUE)),
    Max_Weight        = ifelse(all(is.na(weight_g)), NA, max(weight_g, na.rm = TRUE))
  )

CPUE_BPUE_weighted <- CPUE_BPUE_legacy %>%
  group_by(monitoring_id, species) %>%
  summarise(
    Total_Individuals       = sum(Total_Individuals, na.rm = TRUE),
    Total_Weight            = sum(Total_Weight,      na.rm = TRUE),
    Weighted_CPUE_numerator = sum(CPUE * Total_Effort,  na.rm = TRUE),
    Weighted_BPUE_numerator = sum(BPUE * Total_Effort,  na.rm = TRUE),
    Total_Effort_sum        = sum(Total_Effort,         na.rm = TRUE),
    Mean_Length             = mean(Mean_Length, na.rm = TRUE),
    Min_Length              = ifelse(all(is.na(Min_Length)), NA, min(Min_Length, na.rm = TRUE)),
    Max_Length              = ifelse(all(is.na(Max_Length)), NA, max(Max_Length, na.rm = TRUE)),
    Mean_Weight             = mean(Mean_Weight, na.rm = TRUE),
    Min_Weight              = ifelse(all(is.na(Min_Weight)), NA, min(Min_Weight, na.rm = TRUE)),
    Max_Weight              = ifelse(all(is.na(Max_Weight)), NA, max(Max_Weight, na.rm = TRUE)),
    .groups = "drop") %>%
  ungroup() %>%
  mutate(Total_Effort_sum = ifelse(monitoring_id %in% c("16_0", "10_1", "3_2", "16_2"), 3, 4),
    Weighted_CPUE    = Weighted_CPUE_numerator / Total_Effort_sum,
    Weighted_BPUE    = Weighted_BPUE_numerator / Total_Effort_sum)

species_presence_absence <- Fish_data %>%
  filter(!is.na(species)) %>%
  distinct(monitoring_id, species) %>%
  mutate(presence = 1) %>%
  pivot_wider(
    names_from  = species,
    values_from = presence,
    values_fill = list(presence = 0),
    names_prefix = "Presence_") %>%
  mutate(Predator_Fish_Presence = pmax(Presence_Trout, Presence_Eel, Presence_Catfish))

CPUE_BPUE_weighted_summary <- CPUE_BPUE_weighted %>%
  pivot_wider(
    names_from  = species,
    values_from = c(
      Total_Individuals, Weighted_CPUE, Weighted_BPUE, Total_Weight,
      Mean_Length, Mean_Weight, Weighted_CPUE_numerator, Weighted_BPUE_numerator,
      Total_Effort_sum, Min_Length, Max_Length, Min_Weight, Max_Weight),
    names_sep   = "_", values_fill = list(Total_Individuals = 0, Weighted_CPUE = 0, Weighted_BPUE = 0)) %>%
  mutate(
    Richness  = rowSums(dplyr::select(., starts_with("Total_Individuals_")) > 0),
    Abundance = rowSums(dplyr::select(., starts_with("Total_Individuals_") & !ends_with(c("_Bullies", "_Common_smelt")))))
```


## Combined dataset construction
```{r}
#| label: combine-data
#| include: true

unit_metadata <- dplyr::bind_rows(Monitoring_data %>%
    dplyr::select(parameter, unit), Chemical_data %>%
    dplyr::select(parameter, unit)) %>%
  dplyr::distinct() %>%
  dplyr::rename(Variable = parameter, Unit = unit)

Monitoring_summary <- Monitoring_data %>%
  dplyr::select(-group, -notes, -unit, -monitoring_id) %>%
  tidyr::pivot_wider(names_from  = parameter,
    values_from = value,
    values_fill = list(value = NA)) %>%
  dplyr::mutate(
    dplyr::across(c(Riparian_vegetation, Vegetation_nearby, Overhanging_trees, Erosion, Structure, Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter, Wood_cover ), ~ as.numeric(.)))

Chemical_summary <- Chemical_data %>%
  dplyr::select(-group, -notes, -unit, -site_id) %>%
  tidyr::pivot_wider(names_from  = parameter,
    values_from = value,
    values_fill = list(value = NA)) %>%
  dplyr::mutate(
    dplyr::across(c(Temperature, DO_mgl, DO_percent, Conductivity, Specific_conductivity, pH), ~ as.numeric(.)))

Weed_summary <- Weed_data %>%
  group_by(monitoring_id, weed_type, native_status) %>%
  summarise(Total_Cover = sum(percentage_cover, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from  = c(weed_type, native_status), values_from = Total_Cover, values_fill = 0) %>% 
  dplyr::select(monitoring_id, Emergent_Native, Emergent_Non_Native, Submerged_Non_Native, Turf_Native)

Stone_pile_summary <- Stone_pile_data %>%
  select(-unit, -notes, -site_id)%>%
  pivot_wider(names_from = c(parameter, stone_pile_id), values_from = value, names_sep = ".")

Macroinvertebrates_sum <- Macroinvertebrates %>%
  group_by(monitoring_id, species) %>%
  summarise(Total_amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = c(species), values_from = Total_amount, values_fill = 0) %>%
  mutate(Invertebrates_Richness  = rowSums(dplyr::select(., -monitoring_id) > 0),
    Invertebrates_Abundance = rowSums(dplyr::select(., -monitoring_id)))

Combined_data <- Site_info %>%
  left_join(Monitoring_summary, by = "site_id") %>%
  left_join(Chemical_summary, by = c("monitoring_id")) %>%
  left_join(Weed_summary, by = c("monitoring_id")) %>%
  left_join(Stone_pile_summary, by = c("monitoring_id")) %>%
  left_join(CPUE_BPUE_weighted_summary, by = "monitoring_id") %>%
  left_join(species_presence_absence, by = "monitoring_id") %>%
  left_join(Macroinvertebrates_sum, by = "monitoring_id")

```

## Derived variables
```{r}
#| label: derived-variables
#| include: true

Combined_data <- Combined_data %>%
  dplyr::mutate(date_time = as.POSIXct(date_time, tz = "Pacific/Auckland"),
    Presence_rocks = dplyr::if_else(Cobble > 1 | Boulders > 1, 1, 0),
    Period = factor(dplyr::if_else(monitoring == 0, "Before", "After"),levels = c("Before", "After")),
    Date  = as.Date(date_time),
    Time  = format(date_time, "%H:%M:%S"),
    Year  = lubridate::year(date_time),
    Month = lubridate::month(date_time, label = TRUE),
    Day   = lubridate::day(date_time),
    Season = dplyr::case_when(
      Month %in% c("Dec", "Jan", "Feb") ~ "Summer",
      Month %in% c("Mar", "Apr", "May") ~ "Autumn",
      Month %in% c("Jun", "Jul", "Aug") ~ "Winter",
      Month %in% c("Sep", "Oct", "Nov") ~ "Spring",
      TRUE ~ NA_character_),
    Date_Time_Numeric = as.numeric(date_time),
    Monitoring_label = dplyr::case_when(
      monitoring == 0 ~ "0 Pre/Summer",
      monitoring == 1 ~ "1 Autumn",
      monitoring == 2 ~ "2 Winter",
      monitoring == 3 ~ "3 Spring",
      monitoring == 4 ~ "4 Summer",
      monitoring == 5 ~ "5 Autumn",
      monitoring == 6 ~ "6 Winter",
      TRUE ~ as.character(monitoring)
    ) %>% as.factor(), site_id = as.factor(site_id)) %>%
  dplyr::mutate(
  Depth_5m    = suppressWarnings(as.numeric(Distance_5m)),
  Depth_20m   = suppressWarnings(as.numeric(Distance_20m)),
  Slope_0_5m  = 5  / Depth_5m,    # 5m rise / horizontal distance in m
  Slope_0_20m = 20 / Depth_20m)   # 20m rise / horizontal distance in m
        

habitat_classification <- Combined_data %>%
  filter(monitoring == 0) %>%
  dplyr::select(site_id, DHT, lake,
                Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter,
                Emergent_Native, Emergent_Non_Native, Submerged_Non_Native, Wood_cover) %>%
  pivot_longer(cols = c(Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter,
                  Emergent_Native, Emergent_Non_Native, Submerged_Non_Native, Wood_cover),
    names_to  = "Type", values_to = "Percentage") %>%
  group_by(site_id) %>%
  summarise(Rocky_Percentage  = sum(Percentage[Type %in% c("Bedrock", "Boulders", "Cobble")], na.rm = TRUE),
    Sand_Percentage     = sum(Percentage[Type == "Sand"], na.rm = TRUE),
    Mud_Percentage      = sum(Percentage[Type %in% c("Mud", "Organic_matter")], na.rm = TRUE),
    Emergent_Percentage = sum(Percentage[Type %in% c("Emergent_Native")], na.rm = TRUE),
    Substrate_index_before = sum(
      0.08 * Percentage[Type == "Bedrock"] +
      0.07 * Percentage[Type == "Boulders"] +
      0.06 * Percentage[Type == "Cobble"] +
      0.04 * Percentage[Type == "Gravel"] +
      0.03 * Percentage[Type == "Sand"] +
      0.02 * Percentage[Type == "Organic_matter"] +
      0.01 * Percentage[Type == "Mud"],
      na.rm = TRUE),.groups = "drop") %>%
  mutate(Habitat_Type = case_when(
      Rocky_Percentage    > 25 ~ "Rocky",
      Emergent_Percentage > 25 ~ "Emergent Macrophyte",
      Sand_Percentage   >= Mud_Percentage ~ "Sandy",
      TRUE ~ "Muddy")) %>%
  dplyr::select(site_id, Habitat_Type, Substrate_index_before)

Combined_data <- Combined_data %>%
  left_join(habitat_classification, by = c("site_id"))%>%
  dplyr::mutate(monitoring_int = as.integer(monitoring),
    pair_id = factor(ceiling(as.numeric(as.character(site_id)) / 2)),
    koura_present = as.integer(`Total_Individuals_Kōura` > 0),
    Site_type_detail = factor(dplyr::case_when(
      site == "Stone_pile" ~ "Stone_pile",
      site == "Control"    ~ "Control",
      site == "Reference" & as.numeric(as.character(site_id)) %% 2 == 1 ~ "Reference_1",
      site == "Reference" & as.numeric(as.character(site_id)) %% 2 == 0 ~ "Reference_2",
      TRUE ~ site),
      levels = c("Stone_pile", "Control", "Reference_1", "Reference_2")))

site_area_m2 <- 10 * 15

pile_footprint_pct <- Stone_pile_data %>%
  dplyr::filter(stringr::str_detect(monitoring_id, "_0"), parameter == "Area") %>%
  dplyr::mutate(value_num = suppressWarnings(as.numeric(value)),
                site_id   = as.character(site_id)) %>%
  dplyr::group_by(site_id) %>%
  dplyr::summarise(footprint_pct = sum(value_num, na.rm = TRUE) / site_area_m2 * 100, .groups = "drop")

substrate_updated <- Combined_data %>%
  dplyr::filter(monitoring_int == 0) %>%
  dplyr::distinct(site_id, Bedrock, Boulders, Cobble, Gravel, Sand, Mud, Organic_matter) %>%
  dplyr::mutate(site_id = as.character(site_id)) %>%
  dplyr::left_join(pile_footprint_pct, by = "site_id") %>%
  tidyr::replace_na(list(footprint_pct = 0)) %>%
  dplyr::mutate(
    Boulders_new = Boulders + footprint_pct * 0.20,
    Cobble_new   = Cobble   + footprint_pct * 0.80,
    Sand_new     = Sand     - footprint_pct,
    Substrate_index_after = 0.08 * Bedrock + 0.07 * Boulders_new + 0.06 * Cobble_new +
      0.04 * Gravel + 0.03 * Sand_new + 0.02 * Organic_matter + 0.01 * Mud) %>%
  dplyr::select(site_id, Substrate_index_after)

Combined_data <- Combined_data %>%
  dplyr::mutate(site_id = as.character(site_id)) %>%
  dplyr::left_join(substrate_updated, by = "site_id") %>%
  dplyr::mutate(
    site_id = factor(site_id, levels = as.character(sort(as.numeric(unique(site_id))))),
    Substrate_index_after = dplyr::if_else(
      site == "Stone_pile" & monitoring_int >= 1,
      Substrate_index_after,
      Substrate_index_before))

Combined_data <- Combined_data %>%
  mutate(lunar_phase = lunar.phase(Date, name = FALSE),  # 0-2π radians
    lunar_illumination = (1 - cos(lunar.phase(Date))) / 2,  # 0=new, 1=full
    lunar_name = lunar.phase(Date, name = TRUE))     # "New", "Waxing", "Full", "Waning"

monitoring_start <- as.Date("2025-01-01")
monitoring_end   <- as.Date("2026-08-31")

met_daily <- dplyr::bind_rows(
  readRDS(file.path(raw_data_dir, "met_chunk_2025.rds")),
  readRDS(file.path(raw_data_dir, "met_chunk_2026.rds"))
) |>
  dplyr::filter(
    site_id   == "Rotoiti",
    parameter %in% c("TmpAir", "RadSWD", "PpRain", "WndSpd"),
    !is.na(value)
  ) |>
  dplyr::mutate(Date = as.Date(datetime)) |>
  dplyr::filter(Date >= monitoring_start, Date <= monitoring_end) |>
  dplyr::group_by(Date, parameter) |>
  dplyr::summarise(
    value_mean = mean(value, na.rm = TRUE),
    value_sum  = sum(value,  na.rm = TRUE),   # daily total for rainfall
    .groups    = "drop"
  ) |>
  dplyr::mutate(value_plot = dplyr::if_else(parameter == "PpRain", value_sum, value_mean)) |>
  dplyr::select(Date, parameter, value_plot) |>
  tidyr::pivot_wider(names_from = parameter, values_from = value_plot)

readr::write_csv(met_daily, file.path(der_data_dir, "met_daily_rotoiti.csv"))
cat("Meteorological data extracted:", nrow(met_daily), "days,",
    format(range(met_daily$Date)), "\n")

CTD_buoy <- readRDS(file.path(raw_data_dir, "buoy_raw_data.rds"))

Buoy_data_Rotoiti <- CTD_buoy |>
    dplyr::select(Lake, DateTime, DptSns, TmpWtr, DOconc) |>
    dplyr::filter(Lake == "Rotoiti") |>
    dplyr::rename(Depth = DptSns, Temp = TmpWtr, DO = DOconc) |>
    dplyr::mutate(DateTime = as.POSIXct(DateTime, format = "%Y-%m-%d %H:%M:%S")) |>
    dplyr::distinct(Lake, DateTime, .keep_all = TRUE) |>
    tidyr::drop_na(DateTime, Depth) |>
    mutate(
      Date  = as.Date(DateTime),
      Time  = format(DateTime, "%H:%M:%S"),
      Year  = lubridate::year(DateTime),
      Month = lubridate::month(DateTime, label = TRUE),
      Day   = lubridate::day(DateTime),
      Season = case_when(
        lubridate::month(DateTime) %in% c(12, 1, 2) ~ "Summer",
        lubridate::month(DateTime) %in% c(3, 4, 5)  ~ "Autumn",
        lubridate::month(DateTime) %in% c(6, 7, 8)  ~ "Winter",
        lubridate::month(DateTime) %in% c(9, 10, 11) ~ "Spring",
        TRUE ~ NA_character_),
      Date_Time_Numeric = as.numeric(DateTime),
      Time_of_Day = case_when(
        lubridate::hour(DateTime) >= 6  & lubridate::hour(DateTime) < 12 ~ "Morning",
        lubridate::hour(DateTime) >= 12 & lubridate::hour(DateTime) < 18 ~ "Afternoon",
        lubridate::hour(DateTime) >= 18 & lubridate::hour(DateTime) < 24 ~ "Evening",
        TRUE ~ "Night"))

Buoy_study <- Buoy_data_Rotoiti |>
    dplyr::filter(Date >= as.Date("2024-07-01"),
                  Date <= as.Date("2026-07-31"))

oxycline_daily <- Buoy_study |>
    dplyr::group_by(Date) |>
    dplyr::arrange(Depth, .by_group = TRUE) |>
    dplyr::summarise(
        habitable_depth_m = {
            d <- Depth[DO >= 5 & is.finite(DO)]
            if (length(d) == 0) NA_real_ else max(d, na.rm = TRUE)},
        hypoxia_top_m = {
            d <- Depth[DO < 5 & is.finite(DO)]
            if (length(d) == 0) NA_real_ else min(d, na.rm = TRUE)},
        surface_temp = mean(Temp[Depth < 5], na.rm = TRUE),
        .groups = "drop" ) |>
    dplyr::mutate(full_column = is.na(hypoxia_top_m))

visit_dates <- Combined_data |>
    dplyr::mutate(visit_date = as.Date(date_time)) |>
    dplyr::select(site_id, monitoring, monitoring_id, visit_date)

visit_buoy <- visit_dates |>
    dplyr::rowwise() |>
    dplyr::mutate(buoy_slice = list(oxycline_daily |>
     dplyr::filter(Date >= visit_date - 14, Date <= visit_date + 14)),
        habitable_depth_m = mean(buoy_slice$habitable_depth_m, na.rm = TRUE),
        hypoxia_top_m = mean(buoy_slice$hypoxia_top_m, na.rm = TRUE),
        surface_temp_C = mean(buoy_slice$surface_temp, na.rm = TRUE),
        full_column = all(buoy_slice$full_column, na.rm = TRUE)) |>
    dplyr::select(-buoy_slice) |>
    dplyr::ungroup()

Combined_data <- Combined_data |>
    dplyr::left_join(visit_buoy |>
    dplyr::select(monitoring_id, habitable_depth_m, hypoxia_top_m, surface_temp_C, full_column),
        by = "monitoring_id")

compromised_ids <- c("6_1", "13_1", "14_1", "1_2", "2_2", "19_2", "20_2", "1_3", "2_3", "6_5", "5_6", "6_6")

Combined_data <- Combined_data %>%
  mutate(monitoring_quality = case_when(monitoring_id %in% compromised_ids ~ "compromised", is.na(`Total_Individuals_Kōura`) ~ "missing", TRUE  ~ "complete"))

writexl::write_xlsx(Combined_data, file.path(der_data_dir, "Combined_data.xlsx"))
write.csv(Combined_data, file.path(der_data_dir, "Combined_data.csv"), row.names = FALSE)
write.csv(habitat_classification, file.path(der_data_dir, "habitat_classification.csv"), row.names = FALSE)

all_data <- Combined_data

# Join fetch and wave exposure covariates (from fetch-calc chunk)
site_fetch_cov <- readr::read_csv(file.path(der_data_dir, "site_fetch.csv"),
                                   show_col_types = FALSE) |>
  dplyr::select(site_id, max_eff_fetch_km) |>
  dplyr::mutate(site_id = factor(site_id, levels = levels(all_data$site_id)))

wave_round_cov <- readr::read_csv(file.path(der_data_dir, "wave_exposure_by_round.csv"),
                                   show_col_types = FALSE) |>
  dplyr::select(site_id, round, wave_energy_proxy) |>
  dplyr::mutate(site_id = factor(site_id, levels = levels(all_data$site_id)))

all_data <- all_data |>
  dplyr::left_join(site_fetch_cov, by = "site_id") |>
  dplyr::left_join(wave_round_cov, by = c("site_id", "monitoring_int" = "round")) |>
  dplyr::mutate(
    fetch_z = as.numeric(scale(max_eff_fetch_km)),
    wave_z  = as.numeric(scale(wave_energy_proxy))
  )

```

## Lake conditions heatmap
```{r}
#| label: fig-lake-conditions
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Depth-resolved water temperature (°C, top) and dissolved oxygen (mg/L, bottom) in Lake Rotoiti throughout the study period (February 2025 – July 2026). Data derived from the LimnoTrack continuous monitoring buoy (Moore and McBride, 2026). Vertical green line marks stone pile deployment; dashed vertical lines mark seasonal monitoring rounds. Warm colours in the DO panel (orange–black) indicate hypoxic and anoxic conditions (< 5 mg/L), demonstrating the pronounced summer stratification that compresses oxygenated habitat into the shallow littoral zone."

stone_pile_date <- as.Date("2025-02-14")

monitoring_lines <- visit_dates |>
  dplyr::group_by(monitoring) |>
  dplyr::summarise(date = median(visit_date), .groups = "drop")

buoy_daily <- Buoy_study |>
  dplyr::mutate(
    Date      = as.Date(DateTime),
    Depth_bin = round(Depth)
  ) |>
  dplyr::group_by(Date, Depth_bin) |>
  dplyr::summarise(
    DO_mean   = mean(DO,   na.rm = TRUE),
    Temp_mean = mean(Temp, na.rm = TRUE),
    .groups = "drop"
  )

x_lims <- c(
  max(min(buoy_daily$Date), min(monitoring_lines$date) - 14),
  min(max(buoy_daily$Date), max(monitoring_lines$date) + 14)
)

p_temp <- ggplot(buoy_daily, aes(x = Date, y = Depth_bin, fill = Temp_mean)) +
  geom_tile() +
  geom_vline(data = monitoring_lines, aes(xintercept = date),
             colour = "black", linetype = "solid", linewidth = 1.2) +
  geom_vline(data = monitoring_lines, aes(xintercept = date),
             colour = "white", linetype = "dashed", linewidth = 0.5) +
  geom_text(data = monitoring_lines, aes(x = date, label = paste0("R", monitoring)),
            y = 2, colour = "white", size = 2.8, fontface = "bold",
            hjust = -0.15, inherit.aes = FALSE) +
  geom_vline(xintercept = stone_pile_date,
             colour = "#39FF14", linetype = "solid", linewidth = 1.4) +
  scale_y_reverse(expand = c(0, 0)) +
  scale_x_date(limits = x_lims, expand = c(0, 0)) +
  scale_fill_gradientn(
    colours = c("#00008B", "#4B0082", "#9400D3", "#FF8C00", "#FFD700"),
    values  = scales::rescale(c(10, 14, 17, 21, 25)),
    limits  = c(10, 25),
    oob     = scales::squish,
    na.value = "grey85",
    name    = "Temp\n(°C)",
    breaks  = c(10, 14, 17, 20, 23)
  ) +
  labs(x = NULL, y = "Depth (m)") +
  theme_bw() +
  theme(plot.title = element_text(size = 11))

p_do <- ggplot(buoy_daily, aes(x = Date, y = Depth_bin, fill = DO_mean)) +
  geom_tile() +
  geom_vline(data = monitoring_lines, aes(xintercept = date),
             colour = "black", linetype = "solid", linewidth = 1.2) +
  geom_vline(data = monitoring_lines, aes(xintercept = date),
             colour = "white", linetype = "dashed", linewidth = 0.5) +
  geom_text(data = monitoring_lines, aes(x = date, label = paste0("R", monitoring)),
            y = 2, colour = "white", size = 2.8, fontface = "bold",
            hjust = -0.15, inherit.aes = FALSE) +
  geom_vline(xintercept = stone_pile_date,
             colour = "#39FF14", linetype = "solid", linewidth = 1.4) +
  scale_y_reverse(expand = c(0, 0)) +
  scale_x_date(limits = x_lims, expand = c(0, 0)) +
  scale_fill_gradientn(
    colours = c("grey15", "#8B0000", "#FF4500", "#FFA500", "#ADD8E6", "#00CED1", "#00FA9A"),
    values  = scales::rescale(c(0, 1, 2, 4, 8, 12, 16)),
    limits  = c(0, 16),
    oob     = scales::squish,
    na.value = "grey85",
    name    = "DO\n(mg/L)",
    breaks  = c(0, 4, 8, 12, 16)
  ) +
  labs(x = "Date", y = "Depth (m)") +
  theme_bw() +
  theme(plot.title = element_text(size = 11))

fig_lake_conditions <- p_temp / p_do
fig_lake_conditions

ggsave(file.path(out_dir, "fig-lake-conditions.png"),
       fig_lake_conditions, width = 12, height = 8, dpi = 300)
```

## Meteorological conditions
```{r}
#| label: fig-met-conditions
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Meteorological conditions at Lake Rotoiti during the monitoring period. From top: air temperature (°C), shortwave downwelling radiation (W/m²), daily rainfall (mm), and wind speed (m/s). Vertical lines mark monitoring visits; orange line marks stone pile construction completion (14 February 2025)."

met_plot <- met_daily |>
  tidyr::pivot_longer(cols = c(TmpAir, RadSWD, PpRain, WndSpd),
                      names_to = "parameter", values_to = "value") |>
  dplyr::mutate(parameter = factor(parameter,
    levels = c("TmpAir", "RadSWD", "PpRain", "WndSpd"),
    labels = c("Air temperature (°C)", "Solar radiation (W/m²)",
               "Daily rainfall (mm)", "Wind speed (m/s)")))

p_met <- ggplot(met_plot, aes(x = Date, y = value)) +
  geom_line(colour = "grey40", linewidth = 0.4) +
  geom_vline(data = monitoring_lines, aes(xintercept = date),
             colour = "black", linetype = "dashed", linewidth = 0.5) +
  geom_vline(xintercept = as.Date("2025-02-14"),
             colour = "#D55E00", linewidth = 0.8) +
  facet_wrap(~ parameter, ncol = 1, scales = "free_y") +
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y",
               limits = c(monitoring_start, monitoring_end)) +
  labs(x = NULL, y = NULL) +
  theme(strip.background = element_blank(),
        strip.text = element_text(hjust = 0))

p_met

ggsave(file.path(out_dir, "fig-met-conditions.png"),
       p_met, width = 12, height = 8, dpi = 300)
```

## Wave exposure (fetch analysis)
```{r}
#| label: fetch-calc
#| message: false
#| warning: false

if (!requireNamespace("sf", quietly = TRUE)) install.packages("sf")
library(sf)

lake_poly <- sf::st_read(file.path(raw_data_dir, "Lake Rotoiti.gpkg"), quiet = TRUE) |>
  sf::st_transform(crs = 2193) |>   # NZTM — metres, accurate for NZ
  sf::st_union() |>
  sf::st_buffer(1)                   # 1 m only — just handles floating-point edge cases

lake_boundary <- sf::st_boundary(lake_poly)   # shoreline linestring for crossing detection

# --- 2. Unique site locations ------------------------------------------------
sites_sf <- Site_info |>
  dplyr::distinct(site_id, site, lat, lon) |>
  sf::st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  sf::st_transform(crs = 2193)

# --- 3. Fetch per direction --------------------------------------------------
# Cast a ray from each site outward; find where it FIRST crosses the lake
# boundary after leaving the site (> 50 m away). This correctly stops at
# peninsulas — it does not continue to count water behind land.
directions <- seq(0, 350, by = 10)   # 36 directions, every 10°

calc_fetch <- function(site_pt, lake_poly, lake_bnd, direction_deg, max_dist = 15000) {
  rad     <- direction_deg * pi / 180
  dir_vec <- c(sin(rad), cos(rad))
  coord   <- as.numeric(sf::st_coordinates(site_pt))

  # Guard: at least 30 m of the 5–405 m ray segment must be inside the lake.
  # Starting 5 m ahead avoids boundary floating-point noise.
  near_ray <- sf::st_sfc(
    sf::st_linestring(matrix(
      c(coord + dir_vec * 5, coord + dir_vec * 405),
      ncol = 2, byrow = TRUE
    )),
    crs = 2193
  )
  near_clip <- suppressWarnings(sf::st_intersection(near_ray, lake_poly))
  if (length(near_clip) == 0 || sf::st_is_empty(near_clip[[1]])) return(0)
  if (as.numeric(sf::st_length(near_clip)) < 30) return(0)

  endpt <- coord + dir_vec * max_dist
  ray   <- sf::st_sfc(
    sf::st_linestring(matrix(c(coord, endpt), ncol = 2, byrow = TRUE)),
    crs = 2193
  )
  crossings <- suppressWarnings(sf::st_intersection(ray, lake_bnd))
  if (length(crossings) == 0 || sf::st_is_empty(crossings[[1]])) return(max_dist)

  pts    <- suppressWarnings(sf::st_cast(crossings, "POINT"))
  cxy    <- sf::st_coordinates(pts)
  signed <- apply(cxy, 1, function(p) sum((p - coord) * dir_vec))
  forward <- signed[signed > 20]
  if (length(forward) == 0) return(max_dist)
  min(forward)
}

# Snap each site to the nearest point on the lake boundary.
# This ensures sites that are slightly outside the polygon (coordinate mismatch)
# start exactly on the shoreline, so the guard works correctly for all sites.
sites_snapped <- sites_sf
sf::st_geometry(sites_snapped) <- purrr::map(
  seq_len(nrow(sites_sf)),
  function(i) {
    connector <- sf::st_nearest_points(sites_sf[i, ], lake_boundary)
    sf::st_cast(connector, "POINT")[[2]]
  }
) |> sf::st_sfc(crs = 2193)

message("Calculating fetch: ", nrow(sites_snapped), " sites x ", length(directions), " directions")
fetch_raw <- purrr::map_dfr(seq_len(nrow(sites_snapped)), function(i) {
  site_pt <- sites_snapped[i, ]
  dplyr::tibble(
    site_id   = sites_sf$site_id[i],
    site_type = sites_sf$site[i],
    direction = directions,
    fetch_m   = purrr::map_dbl(directions, function(d) calc_fetch(site_pt, lake_poly, lake_boundary, d))
  )
})

# --- 4. Effective fetch (USACE cosine weighting ±45° in 5° steps) -----------
eff_fetch_at_dir <- function(df, wind_dir) {
  offsets <- seq(-45, 45, by = 5)
  weights <- cos(offsets * pi / 180)
  sectors <- (wind_dir + offsets) %% 360
  f_sect  <- purrr::map_dbl(sectors, function(s) {
    diffs <- pmin(abs(df$direction - s), 360 - abs(df$direction - s))
    df$fetch_m[which.min(diffs)]
  })
  sum(weights * f_sect) / sum(weights)
}

fetch_eff <- fetch_raw |>
  dplyr::group_by(site_id, site_type) |>
  dplyr::group_modify(function(df, key) {
    df$eff_fetch_m <- purrr::map_dbl(df$direction, function(d) eff_fetch_at_dir(df, d))
    df
  }) |>
  dplyr::ungroup()

# Static summary: maximum effective fetch per site (worst-case exposure)
site_exposure <- fetch_eff |>
  dplyr::group_by(site_id, site_type) |>
  dplyr::summarise(
    max_eff_fetch_km   = round(max(eff_fetch_m) / 1000, 2),
    mean_eff_fetch_km  = round(mean(eff_fetch_m) / 1000, 2),
    dominant_dir_deg   = direction[which.max(eff_fetch_m)],
    # Raw fetch (not cosine-averaged) for sanity-check map — arrow should reach the far shore
    max_raw_fetch_km   = round(max(fetch_m) / 1000, 2),
    dominant_raw_dir   = direction[which.max(fetch_m)],
    .groups = "drop"
  )

# 5. Wind direction by monitoring round (14-day pre-visit window) 
met_wind <- dplyr::bind_rows(
  readRDS(file.path(raw_data_dir, "met_chunk_2025.rds")),
  readRDS(file.path(raw_data_dir, "met_chunk_2026.rds"))
) |>
  dplyr::filter(site_id == "Rotoiti",
                parameter %in% c("WndSpd", "WndDir"),
                !is.na(value)) |>
  dplyr::mutate(Date = as.Date(datetime))

round_dates <- Site_info |>
  dplyr::mutate(visit_date = as.Date(date_time)) |>
  dplyr::group_by(monitoring) |>
  dplyr::summarise(start_date = min(visit_date, na.rm = TRUE), .groups = "drop") |>
  dplyr::rename(round = monitoring)

wind_by_round <- purrr::map_dfr(seq_len(nrow(round_dates)), function(i) {
  r_date  <- round_dates$start_date[i]
  window  <- met_wind |>
    dplyr::filter(Date >= r_date - 14, Date < r_date)
  spd     <- window |> dplyr::filter(parameter == "WndSpd") |> dplyr::pull(value)
  dir_deg <- window |> dplyr::filter(parameter == "WndDir") |> dplyr::pull(value)
  dir_rad <- dir_deg * pi / 180
  mean_dir <- atan2(mean(sin(dir_rad), na.rm = TRUE),
                    mean(cos(dir_rad), na.rm = TRUE)) * 180 / pi
  dplyr::tibble(
    round         = round_dates$round[i],
    visit_start   = r_date,
    mean_wind_spd = mean(spd, na.rm = TRUE),
    mean_wind_dir = (mean_dir + 360) %% 360,
    n_obs         = sum(!is.na(dir_deg))
  )
})

# --- 6. Wave energy proxy per site per round ---------------------------------
# Wave energy ~ effective_fetch × wind_speed²  (proportional to Hs²)
wave_exposure <- purrr::map_dfr(seq_len(nrow(wind_by_round)), function(j) {
  r     <- wind_by_round$round[j]
  w_spd <- wind_by_round$mean_wind_spd[j]
  w_dir <- wind_by_round$mean_wind_dir[j]
  fetch_eff |>
    dplyr::group_by(site_id, site_type) |>
    dplyr::group_modify(function(df, key) {
      diffs <- pmin(abs(df$direction - w_dir), 360 - abs(df$direction - w_dir))
      best  <- df$eff_fetch_m[which.min(diffs)]
      dplyr::tibble(
        eff_fetch_m       = best,
        wind_spd_ms       = w_spd,
        wind_dir_deg      = w_dir,
        wave_energy_proxy = best * w_spd^2
      )
    }) |>
    dplyr::mutate(round = r) |>
    dplyr::ungroup()
})

readr::write_csv(site_exposure, file.path(der_data_dir, "site_fetch.csv"))
readr::write_csv(wave_exposure, file.path(der_data_dir, "wave_exposure_by_round.csv"))

cat("Fetch range (max eff fetch):",
    round(range(site_exposure$max_eff_fetch_km), 1), "km\n")
cat("Sites with max fetch > 5 km:",
    sum(site_exposure$max_eff_fetch_km > 5), "\n")
```

```{r}
#| label: fig-fetch-map
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Sanity check: Lake Rotoiti outline with monitoring sites and raw maximum-fetch rays. Each arrow points in the direction of greatest raw fetch; the arrowhead should land on the opposite shoreline. Effective fetch (used in the analysis) is shorter due to cosine-weighting across ±45°."

lake_wgs84 <- sf::st_read(file.path(raw_data_dir, "Lake Rotoiti.gpkg"), quiet = TRUE) |>
  sf::st_transform(crs = 4326) |>
  sf::st_union()   # combine all polygon features so all lake arms are rendered

# Site coordinates with fetch summary in WGS84 (for geom_sf compatibility)
# 1° lat ≈ 111 km; 1° lon ≈ 89 km at latitude -38°
fetch_map_df <- site_exposure |>
  dplyr::left_join(
    Site_info |> dplyr::distinct(site_id, site, lat, lon),
    by = c("site_id", "site_type" = "site")) |>
  dplyr::mutate(
    rad     = dominant_raw_dir * pi / 180,
    lon_end = lon + sin(rad) * max_raw_fetch_km / 89,
    lat_end = lat + cos(rad) * max_raw_fetch_km / 111,
    site_type = factor(site_type, levels = c("Stone_pile", "Control", "Reference"))  )

fetch_map <- ggplot() +
  geom_sf(data = lake_wgs84, fill = "#c6e2f5", colour = "grey50", linewidth = 0.4) +
  geom_segment(
    data  = fetch_map_df,
    aes(x = lon, y = lat, xend = lon_end, yend = lat_end, colour = site_type),
    arrow = arrow(length = unit(0.22, "cm"), type = "closed"),
    linewidth = 0.9, alpha = 0.9) +
  geom_point(
    data  = fetch_map_df,
    aes(x = lon, y = lat, colour = site_type, shape = site_type),
    size = 3, fill = "white", stroke = 1.2) +
  geom_text(
    data  = fetch_map_df,
    aes(x = lon, y = lat, label = site_id),
    vjust = -0.9, size = 2.5, colour = "grey20") +
  scale_colour_manual(values = site_colours, labels = site_labels, name = NULL) +
  scale_shape_manual( values = site_shapes,  labels = site_labels, name = NULL) +
  labs(x = "Longitude", y = "Latitude") +
  coord_sf() 

ggsave(file.path(out_dir, "fig-fetch-map.png"), fetch_map , width = 10, height = 7, dpi = 300)

fetch_map
```

```{r}
#| label: fig-fetch-exposure
#| fig-width: 12
#| fig-height: 5
#| fig-cap: "Wave exposure at monitoring sites. Left: maximum effective fetch per site (km), the worst-case open-water fetch distance accounting for ±45° angular spread. Right: wave energy proxy (effective fetch × wind speed²) per site per monitoring round; colours indicate site type. Each line is one site. Higher values indicate greater potential for wave disturbance."

# Panel A: max effective fetch per site
p_fetch <- site_exposure |>
  dplyr::mutate(
    site_label = paste0("Site ", site_id),
    site_type  = factor(site_type, levels = c("Stone_pile", "Control", "Reference"))
  ) |>
  ggplot(aes(x = reorder(site_label, max_eff_fetch_km),
             y = max_eff_fetch_km, colour = site_type, fill = site_type)) +
  geom_col(width = 0.7, alpha = 0.85) +
  scale_colour_manual(values = site_colours, labels = site_labels, name = NULL) +
  scale_fill_manual(  values = site_fill,    labels = site_labels, name = NULL) +
  labs(x = NULL, y = "Max effective fetch (km)") +
  coord_flip() 

# Panel B: wave energy proxy per site over rounds
p_wave <- wave_exposure |>
  dplyr::left_join(wind_by_round |> dplyr::select(round, visit_start), by = "round") |>
  dplyr::mutate(
    site_type  = factor(site_type, levels = c("Stone_pile", "Control", "Reference")),
    site_label = paste0("Site ", site_id)
  ) |>
  ggplot(aes(x = visit_start, y = wave_energy_proxy / 1e6,
             colour = site_type, group = site_label)) +
  geom_line(linewidth = 0.5, alpha = 0.7) +
  geom_point(size = 1.5, alpha = 0.8) +
  scale_colour_manual(values = site_colours, labels = site_labels, name = NULL) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  labs(x = NULL, y = expression(Wave~energy~proxy~(10^6~m^3~s^{-2}))) 

p_fetch_combined <- p_fetch + p_wave

ggsave(file.path(out_dir, "fig-fetch-exposure.png"), p_fetch_combined, width = 12, height = 5, dpi = 300)

p_fetch_combined


```

## Missing data summary
```{r}
#| label: tbl-missing-data-summary
#| include: true
#| tbl-cap: "Monitoring events excluded due to environmental disturbance."

# Missing data summary
missing_table <- all_data %>%
  filter(monitoring_quality != "complete") %>%
  select(site_id, site, Monitoring_label, monitoring_id, monitoring_quality) %>%
  arrange(site_id, Monitoring_label)

knitr::kable(missing_table,
             col.names = c("Site ID", "Site", "Monitoring label", "Monitoring ID", "Quality"))

```

# Site overview
## Site overview table
```{r}
#| label: tbl-site-overview
#| include: true
#| tbl-cap: "Site overview: baseline physical characteristics by site. Stone pile structural dimensions are reported in Methods and the Stone pile structure figure instead, since they have no Control/Reference equivalent to display alongside."

# Site-level info from all_data — baseline monitoring only
site_info_summary <- all_data %>%
  filter(monitoring == 0) %>%
  group_by(site_id, site, DHT, Habitat_Type,
           Substrate_index_before, Slope_0_20m, max_eff_fetch_km) %>%
  summarise(
    Riparian_veg_pct  = mean(Riparian_vegetation, na.rm = TRUE),
    Overhang_pct      = mean(Overhanging_trees,   na.rm = TRUE),
    Wood_cover_pct    = mean(Wood_cover,           na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(site_id = as.factor(site_id))

# Stone pile summary — mean across 3 piles per modified site at baseline
stone_pile_site_summary <- Stone_pile_data %>%
  filter(str_detect(monitoring_id, "_0")) %>%
  filter(parameter %in% c("Weight", "Volume", "Height_above", 
                           "Distance_to_shore", "Depth_on_R", 
                           "Depth_on_L", "Depth_after", "Area")) %>%
  mutate(value_num = suppressWarnings(as.numeric(value))) %>%
  group_by(site_id, parameter) %>%
  summarise(
    value_num = mean(value_num, na.rm = TRUE),
    n_piles   = n_distinct(stone_pile_id),
    .groups   = "drop") %>%
  pivot_wider(names_from = parameter, values_from = value_num) %>%
  mutate(
    site_id          = as.character(site_id),
    Depth_m          = round((Depth_on_R + Depth_on_L) / 2 / 1000, 2),
    Depth_after_m    = round(Depth_after       / 1000, 2),
    Height_above_m   = round(Height_above       / 1000, 2),
    Distance_shore_m = round(Distance_to_shore / 1000, 2),
    Area_m2          = round(Area,   2),
    Weight_kg        = round(Weight, 0),
    Volume_m3        = round(Volume, 2)) %>%
  select(site_id, Distance_shore_m, Depth_m, Height_above_m, Area_m2, Volume_m3)

# Join and finalise
site_data <- site_info_summary %>%
  left_join(stone_pile_site_summary, by = "site_id") %>%
  arrange(as.numeric(as.character(site_id))) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) %>%
  select(site_id, site, DHT, Substrate_index_before, Slope_0_20m,
         max_eff_fetch_km, Riparian_veg_pct, Overhang_pct, Wood_cover_pct,
         Distance_shore_m, Depth_m, Height_above_m, Area_m2, Volume_m3)

site_overview_display <- site_data %>%
  dplyr::mutate(site = dplyr::recode(site, "Stone_pile" = "Stone pile")) %>%
  dplyr::select(
    Site    = site_id,
    Type    = site,
    Habitat = DHT,
    `Substrate index` = Substrate_index_before,
    `Slope (m/m)`     = Slope_0_20m,
    `Fetch (km)`      = max_eff_fetch_km,
    `Riparian veg. (%)` = Riparian_veg_pct,
    `Overhang (%)` = Overhang_pct,
    `Wood cover (%)` = Wood_cover_pct
  )

write.csv(site_overview_display, file = file.path(out_dir, "tbl-site-overview.csv"), row.names = FALSE)

knitr::kable(site_overview_display, digits = 2,
             caption = "Site physical characteristics at baseline. Sites 1–10 are stone pile (odd) and paired control (even) sites; sites 11–20 are reference sites.")

```

# Descriptive results
## Descriptive context plots
Background/methods-supporting plots — combined into a single figure below
rather than shown individually, since none are primary results.

### Habitat characteristics
```{r}
#| label: fig-habitat-plot
#| include: false
#| fig-cap: "Baseline habitat characteristics (slope, substrate index, riparian vegetation, overhanging tree cover, wood cover), mean ± SE by site type."

Habitat_data <- all_data %>%
  dplyr::select(site_id, site, monitoring_int, Habitat_Type,Slope_0_5m, Slope_0_20m, Substrate_index_before, Substrate_index_after, Riparian_vegetation, Overhanging_trees, Wood_cover) %>%
  tidyr::pivot_longer(cols = c(Slope_0_5m, Slope_0_20m, Substrate_index_before, Substrate_index_after, Riparian_vegetation, Overhanging_trees, Wood_cover), names_to  = "Variable", values_to = "Value") %>%
  dplyr::group_by(site_id, site, Variable) %>%
  dplyr::slice_max(monitoring_int, n = 1, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

Habitat_summary <- Habitat_data %>%
  dplyr::group_by(site, Variable) %>%
  dplyr::summarise(
    mean_v = mean(Value, na.rm = TRUE),
    se     = sd(Value, na.rm = TRUE) / sqrt(sum(!is.na(Value))),
    .groups = "drop")

habitat_var_labels <- c(
  Slope_0_5m             = "Slope (0-5m)",
  Slope_0_20m            = "Slope (0-20m)",
  Substrate_index_before = "Substrate index\n(before)",
  Substrate_index_after  = "Substrate index\n(after)",
  Riparian_vegetation    = "Riparian\nvegetation",
  Overhanging_trees      = "Overhanging\ntrees",
  Wood_cover             = "Wood cover")

Habitat_plot <- ggplot(Habitat_summary, aes(site, mean_v, fill = site)) +
  geom_col(colour = "black") +
  geom_errorbar(aes(ymin = mean_v - se, ymax = mean_v + se), width = 0.25) +
  facet_wrap(~ Variable, scales = "free_y", nrow = 2, labeller = as_labeller(habitat_var_labels)) +
  scale_fill_manual(values = site_fill, labels = site_labels) +
  labs(y = "Mean value (± SE)", x = NULL) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), legend.position = "none")

ggsave(file.path(out_dir, "fig-habitat-plot.png"), Habitat_plot, width = 10, height = 6, dpi = 300)
```

### Stone pile structure
```{r}
#| label: fig-stone-pile-structure
#| include: false
#| fig-cap: "Stone pile structure parameters at each site."

stone_pile_data2 <- site_data %>%
  dplyr::select(site_id, site, Distance_shore_m, Depth_m, Height_above_m, Area_m2, Volume_m3) %>%
  tidyr::pivot_longer(
    cols      = c(Distance_shore_m, Depth_m, Height_above_m, Area_m2, Volume_m3),
    names_to  = "Parameter",
    values_to = "Value") %>%
  dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

parameter_labels <- c(
  Distance_shore_m = "Distance to shore (m)",
  Depth_m          = "Depth (m)",
  Height_above_m   = "Height above lakebed (m)",
  Area_m2          = "Area (m²)",
  Volume_m3        = "Volume (m³)")

stone_pile_structure_plot <- stone_pile_data2 %>%
  dplyr::filter(!is.na(Value)) %>%
  ggplot(aes(site_id, Value)) +
  geom_col(colour = "black", fill = site_fill[["Stone_pile"]]) +
  facet_wrap(~ Parameter, scales = "free_y", nrow = 2, labeller = as_labeller(parameter_labels)) +
  labs(y = "Value", x = "Stone pile site")

ggsave(file.path(out_dir, "fig-stone-pile-structure.png"), stone_pile_structure_plot, width = 8, height = 6, dpi = 300)



# Overall means across sites, used inline in the Methods text
mean_dist_shore <- mean(stone_pile_data2$Value[stone_pile_data2$Parameter == "Distance_shore_m"], na.rm = TRUE)
mean_depth      <- mean(stone_pile_data2$Value[stone_pile_data2$Parameter == "Depth_m"], na.rm = TRUE)
mean_area       <- mean(stone_pile_data2$Value[stone_pile_data2$Parameter == "Area_m2"], na.rm = TRUE)
```

### Chemical conditions over time
```{r}
#| label: fig-chemical-plot
#| include: false
#| fig-cap: "Chemical conditions (temperature, dissolved oxygen, conductivity, pH) over the monitoring period, coloured by site type."

Chemical_data2 <- all_data %>%
  dplyr::select(monitoring_id, monitoring_int, site_id, site, Monitoring_label, Habitat_Type, DO_mgl, DO_percent, Conductivity, Specific_conductivity, pH, Temperature) %>%
  dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference"))) %>%
  tidyr::pivot_longer(cols = c(Temperature, DO_mgl, DO_percent, Conductivity, Specific_conductivity, pH), names_to = "Variable", values_to = "Values")

chemical_var_labels <- c(
  Temperature           = "Temperature (°C)",
  DO_mgl                = "Dissolved oxygen\n(mg/L)",
  DO_percent            = "Dissolved oxygen\n(%)",
  Conductivity          = "Conductivity\n(µS/cm)",
  Specific_conductivity = "Specific conductivity\n(µS/cm)",
  pH                    = "pH")

Chemical_plot <- ggplot(Chemical_data2, aes(monitoring_int, Values, colour = site)) +
  geom_vline(xintercept = 0.25, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
  geom_point(alpha = 0.5, position = position_jitter(width = 0.1)) +
  geom_smooth(aes(group = site), se = FALSE) +
  facet_wrap(~ Variable, scales = "free", nrow = 2, labeller = as_labeller(chemical_var_labels)) +
  scale_colour_manual(values = site_colours, labels = site_labels) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels) +
  labs(y = "Value", x = "Monitoring", colour = "Site type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

ggsave(file.path(out_dir, "fig-chemical-plot.png"), Chemical_plot, width = 8, height = 6,  dpi = 300)
```

### All species trends
```{r}
#| label: fig-species-trends
#| include: false
#| fig-cap: "Mean catch per monitoring round (± 95% CI) by site type for all monitored species."

# Species to plot and their display names
species_vec_trends <- c("Catfish", "Kōaro", "Goldfish")

species_labels <- c("Kōura" = "Kōura", "Catfish" = "Catfish", "Kōaro" = "Kōaro", "Goldfish" = "Goldfish")

# Function to build one trend plot per species
make_trend_plot <- function(sp) {

  col_name <- paste0("Weighted_CPUE_", sp)
  if (!col_name %in% names(all_data)) return(NULL)

  summary_df <- all_data %>%
    group_by(monitoring_int, site) %>%
    summarise(
      mean_n = mean(.data[[col_name]], na.rm = TRUE),
      se     = sd(.data[[col_name]], na.rm = TRUE) /
               sqrt(sum(!is.na(.data[[col_name]]))),
      n      = sum(!is.na(.data[[col_name]])),
      ci95   = qt(0.975, df = pmax(n - 1, 1)) * se,
      .groups = "drop") %>%
    mutate(
      site   = factor(site, levels = c("Stone_pile", "Control", "Reference")),
      ci_low  = pmax(mean_n - ci95, 0),
      ci_high = mean_n + ci95)

  ggplot(summary_df,
      aes(monitoring_int, mean_n, colour = site, shape = site, fill = site)) +
    geom_vline(xintercept = 0.25, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
    annotate("text", x = -0.15, y = Inf, label = "Before", vjust = 1.5,
             hjust = 0.5, size = 2.8, colour = "grey40") +
    annotate("text", x = 3.5,   y = Inf, label = "After",  vjust = 1.5,
             hjust = 0.5, size = 2.8, colour = "grey40") +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high), position = position_dodge(width = 0.4), width = 0.25, linewidth = 0.4, colour = "grey40") +
    geom_line(aes(group = site), position = position_dodge(width = 0.4), linewidth = 0.6) +
    geom_point(colour = "black", size = 2.5, stroke = 0.4, position = position_dodge(width = 0.4)) +
    scale_colour_manual(values = site_colours, labels = site_labels) +
    scale_fill_manual(values = site_fill, labels = site_labels) +
    scale_shape_manual(values = site_shapes, labels = site_labels) +
    scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
    scale_y_continuous(trans = "log1p") +
    labs(title  = species_labels[[sp]], x = NULL, y = "Mean CPUE \n(± 95% CI, log scale)", colour = "Site type", shape = "Site type", fill = "Site type") +
    theme(axis.text.x     = element_text(size = 7,),
      axis.text.y     = element_text(size = 8),
      axis.title.y    = element_text(size = 9),
      plot.title      = element_text(size = 10, face = "bold"),
      legend.position = "none")   # collect legend from patchwork
}

# Generate all plots, drop NULLs
trend_plots <- purrr::map(species_vec_trends, make_trend_plot) %>%
  purrr::set_names(species_vec_trends) %>%
  purrr::compact()

combined_trends <- patchwork::wrap_plots(trend_plots, ncol = 1) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(theme= theme(legend.position = "right")) & theme(legend.position = "right")

# Save combined
ggsave(file.path(out_dir, "fig-all-species-trends.png"), combined_trends, width = 6, height = 6, dpi = 300)

combined_trends

# Also save individual plots if needed
#purrr::iwalk(trend_plots, ~ ggsave(filename = file.path(out_dir, paste0("fig-trend-", .y, ".png")), plot= .x, width= 8, height = 4, units = "in", dpi = 300))

```

### Kōura and catfish combined trends
```{r}
#| label: fig-cpue-trends
#| fig-width: 6
#| fig-height: 5
#| dpi: 300
#| fig-cap: "Mean CPUE per monitoring round (± 95 % CI, log scale) by site type for kōura (top) and catfish (bottom). Vertical dashed line distinguishes the pre-construction period (Before) from post-construction monitoring (After)."

make_cpue_panel <- function(species_col, y_label) {
  all_data %>%
    group_by(monitoring_int, site) %>%
    summarise(
      mean_n  = mean(.data[[species_col]], na.rm = TRUE),
      se      = sd(.data[[species_col]],  na.rm = TRUE) /
                sqrt(sum(!is.na(.data[[species_col]]))),
      n       = sum(!is.na(.data[[species_col]])),
      ci95    = qt(0.975, df = pmax(n - 1, 1)) * se,
      .groups = "drop") %>%
    mutate(
      site    = factor(site, levels = c("Stone_pile", "Control", "Reference")),
      ci_low  = pmax(mean_n - ci95, 0),
      ci_high = mean_n + ci95) %>%
    ggplot(aes(monitoring_int, mean_n, colour = site, shape = site, fill = site)) +
    geom_vline(xintercept = 0.25, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
    annotate("text", x = -0.15, y = Inf, label = "Before", vjust = 1.5,
             hjust = 0.5, size = 3, colour = "grey40") +
    annotate("text", x = 3.5,   y = Inf, label = "After",  vjust = 1.5,
             hjust = 0.5, size = 3, colour = "grey40") +
    geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                  position = position_dodge(width = 0.4),
                  width = 0.25, linewidth = 0.4, colour = "grey40") +
    geom_line(aes(group = site),
              position = position_dodge(width = 0.4), linewidth = 0.6) +
    geom_point(colour = "black", size = 2.5, stroke = 0.4,
               position = position_dodge(width = 0.4)) +
    scale_colour_manual(values = site_colours, labels = site_labels) +
    scale_fill_manual(values   = site_fill,    labels = site_labels) +
    scale_shape_manual(values  = site_shapes,  labels = site_labels) +
    scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
    scale_y_continuous(trans = "log1p") +
    labs(x = NULL, y = y_label,
         colour = "Site type", shape = "Site type", fill = "Site type") +
    theme(
      axis.text.x     = element_text(size = 9),
      axis.text.y     = element_text(size = 9),
      axis.title.y    = element_text(size = 10),
      legend.position = "none")
}

p_koura   <- make_cpue_panel("Weighted_CPUE_Kōura",   "Mean kōura CPUE\n(± 95% CI, log scale)")
p_catfish <- make_cpue_panel("Weighted_CPUE_Catfish", "Mean catfish CPUE\n(± 95% CI, log scale)")

cpue_combined <- (p_koura / p_catfish) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "right")

ggsave(file.path(out_dir, "fig-cpue-trends.png"), cpue_combined, width = 6, height = 5, dpi = 300)

cpue_combined
```

### Combined figure
```{r}
#| label: fig-descriptive-context
#| fig-width: 14
#| fig-height: 12
#| dpi: 300
#| fig-cap: "Descriptive context plots, combined. Left column (top to bottom): baseline habitat characteristics (mean ± SE by site type), stone pile structure dimensions (5 stone pile sites), chemical conditions over time (mean ± smoothed trend by site type). Right column: mean catch per monitoring round by site type, for catfish, kōaro, and goldfish. Site-type colours are shared across all panels — see the legend on the right. Background/methods-supporting figures, not primary results — see Results for the BACI, trend, and convergence models."

left_column <- Habitat_plot / stone_pile_structure_plot / Chemical_plot

descriptive_context_plot <- (left_column | combined_trends) +
  patchwork::plot_layout(widths = c(1.3, 1)) +
  patchwork::plot_annotation(tag_levels = "A")

ggsave(file.path(out_dir, "fig-descriptive-context.png"), descriptive_context_plot, width = 14, height = 12, dpi = 300)

descriptive_context_plot
```

### Kōura trend plot
```{r}
#| label: fig-koura-trends
#| fig-width: 6
#| fig-height: 3
#| dpi: 300
#| fig-cap: "Mean kōura CPUE per monitoring round (± 95% CI, log scale) by site type. Vertical dashed line distinguishes the pre-construction period (Before) from post-construction monitoring (After)."

koura_summary <- all_data %>%
  group_by(monitoring_int, site) %>%
  summarise(
    mean_n  = mean(`Weighted_CPUE_Kōura`, na.rm = TRUE),
    se      = sd(`Weighted_CPUE_Kōura`, na.rm = TRUE) /
              sqrt(sum(!is.na(`Weighted_CPUE_Kōura`))),
    n       = sum(!is.na(`Weighted_CPUE_Kōura`)),
    ci95    = qt(0.975, df = pmax(n - 1, 1)) * se,
    .groups = "drop") %>%
  mutate(
    site    = factor(site, levels = c("Stone_pile", "Control", "Reference")),
    ci_low  = pmax(mean_n - ci95, 0),
    ci_high = mean_n + ci95)

koura_trend <- ggplot(koura_summary,
    aes(monitoring_int, mean_n, colour = site, shape = site, fill = site)) +
  # Before/After separator
  geom_vline(xintercept = 0.25, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
  annotate("text", x = -0.15, y = Inf, label = "Before", vjust = 1.5,
           hjust = 0.5, size = 3, colour = "grey40") +
  annotate("text", x = 3.5, y = Inf, label = "After",  vjust = 1.5,
           hjust = 0.5, size = 3, colour = "grey40") +
  # Data layers
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high),
                position = position_dodge(width = 0.4),
                width = 0.25, linewidth = 0.4, colour = "grey40") +
  geom_line(aes(group = site),
            position = position_dodge(width = 0.4), linewidth = 0.6) +
  geom_point(colour = "black", size = 2.5, stroke = 0.4,
             position = position_dodge(width = 0.4)) +
  scale_colour_manual(values = site_colours, labels = site_labels) +
  scale_fill_manual(values   = site_fill,    labels = site_labels) +
  scale_shape_manual(values  = site_shapes,  labels = site_labels) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
  scale_y_continuous(trans = "log1p") +
  labs(x = NULL, y = "Mean kōura CPUE\n(± 95% CI, log scale)",
       colour = "Site type", shape = "Site type", fill = "Site type") +
  theme(
    axis.text.x     = element_text(size = 9),
    axis.text.y     = element_text(size = 9),
    axis.title.y    = element_text(size = 10),
    legend.position = "right")

ggsave(file.path(out_dir, "fig-koura-trends.png"), koura_trend, width = 6, height = 3, dpi = 300)

koura_trend
```

### Kōura CPUE plot per site
```{r}
#| label: fig-koura-cpue
#| include: true
#| fig-width: 8
#| fig-height: 7
#| fig-cap: "Kōura CPUE (log scale) over time at each individual site, faceted by site ID. Point shape and smoother colour indicate site type."

site_id_order <- c(1, 2, 11, 12, 3, 4, 13, 14, 5, 6, 15, 16, 7, 8, 17, 18, 9, 10, 19, 20)

koura_cpue_plot <- all_data %>%
  dplyr::mutate(
    site    = factor(site, levels = c("Stone_pile", "Control", "Reference")),
    site_id = factor(as.numeric(as.character(site_id)), levels = site_id_order)) %>%
  ggplot(aes(monitoring_int, `Weighted_CPUE_Kōura`, shape = site, fill = site)) +
  geom_vline(xintercept = 0.25, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
  geom_smooth(se = FALSE, aes(group = site, col = site)) +
  geom_point(size = 2, col = "black") +
  facet_wrap(~site_id, ncol = 4) +
  scale_colour_manual(values = site_colours, labels = site_labels) +
  scale_fill_manual(  values = site_fill,    labels = site_labels) +
  scale_shape_manual( values = site_shapes,  labels = site_labels) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
  scale_y_continuous(trans = "log1p") +
  labs(y = "Kōura CPUE (log scale)", x = "Monitoring",
       col = "Site type", shape = "Site type", fill = "Site type") +
  theme(axis.text.x     = element_text(angle = 45, hjust = 1, size = 7),
        legend.position = "right")

ggsave(file.path(out_dir, "fig-koura-cpue.png"), koura_cpue_plot, width = 10, height = 7, dpi = 300)

koura_cpue_plot
```

### Kōura:Catfish log-ratio over time
```{r}
#| label: fig-koura-catfish-ratio
#| fig-width: 7
#| fig-height: 4
#| dpi: 300
#| fig-cap: "Kōura:Catfish log-ratio [log(Kōura CPUE + 1) − log(Catfish CPUE + 1)] over monitoring rounds by site type (mean ± 95% CI). Values above zero indicate kōura-dominated catches; values below zero indicate catfish-dominated catches. The +1 offset handles zero catches."

ratio_data <- all_data %>%
  dplyr::mutate(
    log_ratio = log(`Weighted_CPUE_Kōura` + 1) - log(Weighted_CPUE_Catfish + 1),
    site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

ratio_summary <- ratio_data %>%
  dplyr::group_by(monitoring_int, site) %>%
  dplyr::summarise(
    mean_ratio = mean(log_ratio, na.rm = TRUE),
    se         = sd(log_ratio, na.rm = TRUE) / sqrt(sum(!is.na(log_ratio))),
    n          = sum(!is.na(log_ratio)),
    ci95       = qt(0.975, df = pmax(n - 1, 1)) * se,
    ci_low     = mean_ratio - ci95,
    ci_high    = mean_ratio + ci95,
    .groups    = "drop")

ratio_plot <- ggplot(ratio_summary,
    aes(monitoring_int, mean_ratio, colour = site, fill = site, shape = site)) +
  geom_vline(xintercept = 0.25, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
  annotate("text", x = -0.15, y = Inf, label = "Before", vjust = 1.5,
           hjust = 0.5, size = 3, colour = "grey40") +
  annotate("text", x = 3.5,   y = Inf, label = "After",  vjust = 1.5,
           hjust = 0.5, size = 3, colour = "grey40") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_ribbon(aes(ymin = ci_low, ymax = ci_high), alpha = 0.15, colour = NA,
              position = position_dodge(width = 0.4)) +
  geom_line(aes(group = site), position = position_dodge(width = 0.4), linewidth = 0.7) +
  geom_point(colour = "black", size = 2.5, stroke = 0.4,
             position = position_dodge(width = 0.4)) +
  scale_colour_manual(values = site_colours, labels = site_labels) +
  scale_fill_manual(  values = site_fill,    labels = site_labels) +
  scale_shape_manual( values = site_shapes,  labels = site_labels) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
  labs(x = NULL,
       y = "log(Kōura CPUE + 1) − log(Catfish CPUE + 1)",
       colour = "Site type", fill = "Site type", shape = "Site type") +
  theme(axis.text.x = element_text(size = 9), legend.position = "right")

ggsave(file.path(out_dir, "fig-koura-catfish-ratio.png"), ratio_plot,
       width = 7, height = 4, dpi = 300)

ratio_plot
```

### Kōura Effect plot
```{r}
#| label: fig-koura-effect
#| fig-width: 9
#| fig-height: 5
#| dpi: 300
#| fig-cap: "Kōura CPUE at individual site pairs across all monitoring rounds. Each panel shows one stone pile/control pair (sites 1–10) or reference pair (sites 11–20). Colour and shape indicate site type. Vertical dotted line marks stone pile construction. y-axis on log(1+x) scale."

# Helper: "sites X & Y" label from site_ids within a pair
pair_site_label <- function(df, base_label_col) {
  df %>%
    dplyr::group_by(across(all_of(base_label_col))) %>%
    dplyr::mutate(
      site_ids   = paste(sort(unique(as.integer(as.character(site_id)))),
                         collapse = " & "),
      pair_label = paste0("Sites ", site_ids)) %>%
    dplyr::ungroup()
}

# Stone pile / control pairs
sp_ctrl_eff <- all_data %>%
  dplyr::filter(site %in% c("Stone_pile", "Control")) %>%
  dplyr::mutate(
    pair_base = paste("Pair", pair_id),
    site_eff  = dplyr::recode(as.character(Site_type_detail),
                  "Stone_pile" = "Stone pile", "Control" = "Control")) %>%
  pair_site_label("pair_base")

# Reference pairs — each pair_id gets its own facet
ref_eff <- all_data %>%
  dplyr::filter(site == "Reference") %>%
  dplyr::mutate(
    ref_rank  = as.integer(factor(as.numeric(as.character(pair_id)))),
    pair_base = paste("Reference pair", ref_rank),
    site_eff  = dplyr::recode(as.character(Site_type_detail),
                  "Reference_1" = "Reference1", "Reference_2" = "Reference2")) %>%
  pair_site_label("pair_base")

# Ordered levels
sp_levels  <- sp_ctrl_eff %>%
  dplyr::distinct(pair_base, pair_label) %>%
  dplyr::arrange(pair_base) %>%
  dplyr::pull(pair_label)
ref_levels <- ref_eff %>%
  dplyr::distinct(pair_base, pair_label) %>%
  dplyr::arrange(pair_base) %>%
  dplyr::pull(pair_label)

effect_data <- dplyr::bind_rows(
  dplyr::select(sp_ctrl_eff, monitoring_int, Weighted_CPUE_Kōura,
                pair_label, site_eff, site_id),
  dplyr::select(ref_eff,     monitoring_int, Weighted_CPUE_Kōura,
                pair_label, site_eff, site_id)) %>%
  dplyr::mutate(
    pair_label = factor(pair_label, levels = c(sp_levels, ref_levels)),
    site_eff   = factor(site_eff,
                        levels = c("Stone pile","Control","Reference1","Reference2")))

koura_effect <- ggplot(effect_data,
    aes(x = monitoring_int, y = Weighted_CPUE_Kōura,
        colour = site_eff, fill = site_eff, shape = site_eff,
        group  = site_id)) +
  geom_vline(xintercept = 0.5, linetype = "dotted",
             colour = "grey50", linewidth = 0.6) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5, col="black") +
  facet_wrap(~ pair_label, ncol = 2) +
  scale_colour_manual(values = site_colours_effect, name = "Site type",
                      labels = c("Stone pile","Control","Reference 1","Reference 2")) +
  scale_fill_manual(  values = site_colours_effect, name = "Site type",
                      labels = c("Stone pile","Control","Reference 1","Reference 2")) +
  scale_shape_manual( values = site_shapes_effect,  name = "Site type",
                      labels = c("Stone pile","Control","Reference 1","Reference 2")) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels) +
  scale_y_continuous(trans = "log1p") +
  labs(x = NULL, y = "Kōura CPUE\n(log scale)")

ggsave(file.path(out_dir, "fig-koura-effect.png"), koura_effect, width = 9, height = 9, dpi = 300)

koura_effect
```

## Combined environmental and fish summary table
```{r}
#| label: tbl-env-fish-summary
#| tbl-cap: Distribution of environmental and biotic variables measured at sampling sites. Summaries are grouped by site (ID) across repeated sampling events; an overall ‘All sites’ summary is also provided

group_var <- "site_id" 
var_order_exact <- c("Substrate index","Riparian vegetation","Overhanging trees","Wood cover","Emergent vegetation","Submerged vegetation", "Temperature","Dissolved oxygen","Specific conductivity","pH","Presence Kōura", "Presence Catfish","Presence Eel","Presence Goldfish","Presence Common smelt", "Presence Kōaro","Presence Trout")

ci95 <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  if (n < 2) return(c(NA_real_, NA_real_))
  se <- stats::sd(x) / sqrt(n)
  tcrit <- stats::qt(0.975, df = n - 1)
  m <- mean(x)
  c(m - tcrit * se, m + tcrit * se)
}

site_lookup <- all_data %>%
  dplyr::select(site_id, site) %>%
  dplyr::distinct() %>%
  dplyr::mutate(site_id   = as.character(site_id), site = as.character(site))

unit_map <- dplyr::tibble(Variable = c("Substrate_index_before","Riparian_vegetation","Overhanging_trees","Wood_cover", "Temperature","DO_mgl","Specific_conductivity","pH"))

Env_data <- all_data %>%
  dplyr::mutate(
    Emergent_vegetation  = Emergent_Native + Emergent_Non_Native,
    Submerged_vegetation = Submerged_Non_Native) %>%
  dplyr::select(
    dplyr::all_of(group_var),
    Substrate_index_before, Riparian_vegetation, Overhanging_trees, Wood_cover,
    Emergent_vegetation, Submerged_vegetation,
    Temperature, DO_mgl, Specific_conductivity, pH ) %>%
  tidyr::pivot_longer(
    cols = -dplyr::all_of(group_var),
    names_to = "Variable",
    values_to = "Value")

Env_summary_by_group <- Env_data %>%
  dplyr::group_by(dplyr::across(dplyr::all_of(group_var)), Variable) %>%
  dplyr::summarise(
    n       = sum(!is.na(Value)),
    Mean    = mean(Value, na.rm = TRUE),
    Median  = median(Value, na.rm = TRUE),
    Min     = min(Value, na.rm = TRUE),
    Max     = max(Value, na.rm = TRUE),
    CI_low  = ci95(Value)[1],
    CI_high = ci95(Value)[2],
    .groups = "drop") %>%
  dplyr::rename(Group = !!group_var) %>%
  dplyr::mutate(Group = as.character(Group))

Env_summary_all <- Env_data %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    Group   = "All sites",
    n       = sum(!is.na(Value)),
    Mean    = mean(Value, na.rm = TRUE),
    Median  = median(Value, na.rm = TRUE),
    Min     = min(Value, na.rm = TRUE),
    Max     = max(Value, na.rm = TRUE),
    CI_low  = ci95(Value)[1],
    CI_high = ci95(Value)[2],
    .groups = "drop") %>%
  dplyr::mutate(Group = as.character(Group))

Env_summary_table <- dplyr::bind_rows(Env_summary_by_group, Env_summary_all)

# ---- Fish presence data (long) ----
Fish_data_long <- all_data %>%
  dplyr::select(
    dplyr::all_of(group_var),
    Presence_Kōura, Presence_Catfish, Presence_Eel, Presence_Goldfish, Presence_Common_smelt, Presence_Kōaro, Presence_Trout ) %>%
  tidyr::pivot_longer(
    cols = -dplyr::all_of(group_var),
    names_to = "Variable",
    values_to = "Presence")

Fish_summary_by_group <- Fish_data_long %>%
  dplyr::group_by(dplyr::across(dplyr::all_of(group_var)), Variable) %>%
  dplyr::summarise(
    n       = sum(!is.na(Presence)),
    Mean    = mean(Presence, na.rm = TRUE),
    Median  = median(Presence, na.rm = TRUE),
    Min     = min(Presence, na.rm = TRUE),
    Max     = max(Presence, na.rm = TRUE),
    CI_low  = ci95(Presence)[1],
    CI_high = ci95(Presence)[2],
    .groups = "drop") %>%
  dplyr::rename(Group = !!group_var) %>%
  dplyr::mutate(Group = as.character(Group))

Fish_summary_all <- Fish_data_long %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    Group   = "All sites",
    n       = sum(!is.na(Presence)),
    Mean    = mean(Presence, na.rm = TRUE),
    Median  = median(Presence, na.rm = TRUE),
    Min     = min(Presence, na.rm = TRUE),
    Max     = max(Presence, na.rm = TRUE),
    CI_low  = ci95(Presence)[1],
    CI_high = ci95(Presence)[2],
    .groups = "drop") %>%
  dplyr::mutate(Group = as.character(Group))

Fish_summary_table <- dplyr::bind_rows(Fish_summary_by_group, Fish_summary_all)

# ---- Combine + attach units + pretty variable names + exact ordering ----
EnvBio_summary_table <- dplyr::bind_rows(Env_summary_table, Fish_summary_table) %>%
  dplyr::mutate(Group = as.character(Group)) %>%
  dplyr::left_join(
    site_lookup,
    by = c("Group" = "site_id")) %>%
  dplyr::left_join(unit_metadata, by = "Variable") %>%
  dplyr::mutate(
    Variable = dplyr::recode(
      Variable,
      Substrate_index_before       = "Substrate index",
      Riparian_vegetation   = "Riparian vegetation",
      Overhanging_trees     = "Overhanging trees",
      Wood_cover            = "Wood cover",
      Emergent_vegetation   = "Emergent vegetation",
      Submerged_vegetation  = "Submerged vegetation",
      Temperature           = "Temperature",
      DO_mgl                = "Dissolved oxygen",
      Specific_conductivity = "Specific conductivity",
      pH                    = "pH",
      Presence_Kōura        = "Presence Kōura",
      Presence_Catfish      = "Presence Catfish",
      Presence_Eel          = "Presence Eel",
      Presence_Goldfish     = "Presence Goldfish",
      Presence_Common_smelt = "Presence Common smelt",
      Presence_Kōaro        = "Presence Kōaro",
      Presence_Trout        = "Presence Trout" ),
    Unit = dplyr::coalesce(Unit, "")) %>%
  dplyr::mutate(
    Variable = factor(as.character(Variable), levels = var_order_exact),
    Group    = as.character(Group)) %>%
  dplyr::select(Group, Variable, Unit, n, Mean, Median, Min, Max, CI_low, CI_high) %>%
  dplyr::arrange(Variable, dplyr::desc(Group == "All sites"), Group)

# Optional: if you want Group sorted numerically when Group is ID
if (group_var == "site_id") {
  EnvBio_summary_table <- EnvBio_summary_table %>%
    dplyr::mutate(
      Group_num = suppressWarnings(as.numeric(Group))
    ) %>%
    dplyr::arrange(
      Variable,
      dplyr::desc(Group == "All sites"),
      Group_num,
      Group
    ) %>%
    dplyr::select(-Group_num)
}

write.csv(EnvBio_summary_table, file = file.path(der_data_dir, "EnvBio_summary_table.csv"), row.names = FALSE)

DT::datatable(
  EnvBio_summary_table,
  filter = "top",
  options = list(
    pageLength = 20,
    dom = 'Bfrtip',
    buttons = c('csv', 'excel'),
    columnDefs = list(list(className = 'dt-center', targets = 3:9))
  ),
  extensions = 'Buttons',
  rownames = FALSE,
  colnames = c("Site", "Variable", "Unit", "n", "Mean", "Median", "Min", "Max", "CI low", "CI high")
) |>
  DT::formatRound(columns = c("Mean","Median","Min","Max","CI_low","CI_high"), digits = 2)

```


# Statistical analysis
*Baseline equivalence*
RQ1 — Were stone pile sites equivalent to control sites, and reference sites equivalent to each other, before stone pile installation?

  Descriptive:
    tbl-baseline-combined — Wilcoxon tests across all site-level variables (water quality,
    substrate, CPUE, BPUE); documents overall site comparability and the starting gap
    between stone pile and reference sites

  Inferential:
  RQ1a — Treatment sites (Stone_pile vs Control):
    M1aak - Abundance, Kōura,   Stone_pile vs Control, round 0, Tweedie
    M1aac - Abundance, Catfish, Stone_pile vs Control, round 0, Tweedie
    M1abk - Biomass,   Kōura,   Stone_pile vs Control, round 0, Tweedie
    M1abc - Biomass,   Catfish, Stone_pile vs Control, round 0, Tweedie

  RQ1b — Reference sites (Ref 1 vs Ref 2):
    M1bak - Abundance, Kōura,   Ref 1 vs Ref 2, round 0, Tweedie
    M1bac - Abundance, Catfish, Ref 1 vs Ref 2, round 0, Tweedie
    M1bbk - Biomass,   Kōura,   Ref 1 vs Ref 2, round 0, Tweedie
    M1bbc - Biomass,   Catfish, Ref 1 vs Ref 2, round 0, Tweedie

*Construction impact relative to control sites*
RQ2 — Did kōura and catfish abundance and biomass change at stone pile sites relative to paired control sites after installation?

  RQ2a — Step-change effect (BACI):
    M2aok - Occupancy, Kōura,   Stone_pile vs Control, full timeline, Binomial
    M2aoc - Occupancy, Catfish, Stone_pile vs Control, full timeline, Binomial
    M2aak - Abundance, Kōura,   Stone_pile vs Control, full timeline, Tweedie
    M2aac - Abundance, Catfish, Stone_pile vs Control, full timeline, Tweedie
    M2abk - Biomass,   Kōura,   Stone_pile vs Control, full timeline, Tweedie
    M2abc - Biomass,   Catfish, Stone_pile vs Control, full timeline, Tweedie

  RQ2b — After-construction trajectory:
    M2bok - Occupancy, Kōura,   Stone_pile vs Control, round 1-6, Binomial
    M2boc - Occupancy, Catfish, Stone_pile vs Control, round 1-6, Binomial
    M2bak - Abundance, Kōura,   Stone_pile vs Control, round 1-6, Tweedie
    M2bac - Abundance, Catfish, Stone_pile vs Control, round 1-6, Tweedie
    M2bbk - Biomass,   Kōura,   Stone_pile vs Control, round 1-6, Tweedie
    M2bbc - Biomass,   Catfish, Stone_pile vs Control, round 1-6, Tweedie

*Recovery toward reference conditions*
RQ3 — Did stone pile sites converge toward reference site densities over time, and did reference sites remain equivalent to each other?

  RQ3a — Convergence (Stone_pile vs Reference):
    M3aak - Abundance, Kōura,   Stone_pile vs Reference, round 1-6, Tweedie
    M3aac - Abundance, Catfish, Stone_pile vs Reference, round 1-6, Tweedie
    M3abk - Biomass,   Kōura,   Stone_pile vs Reference, round 1-6, Tweedie
    M3abc - Biomass,   Catfish, Stone_pile vs Reference, round 1-6, Tweedie

  RQ3b — Reference stability (Ref 1 vs Ref 2):
    M3bak - Abundance, Kōura,   Ref 1 vs Ref 2, round 1-6, Tweedie
    M3bac - Abundance, Catfish, Ref 1 vs Ref 2, round 1-6, Tweedie
    M3bbk - Biomass,   Kōura,   Ref 1 vs Ref 2, round 1-6, Tweedie
    M3bbc - Biomass,   Catfish, Ref 1 vs Ref 2, round 1-6, Tweedie

## Baseline equivalence — Were sites comparable before construction?
```{r}
#| label: tbl-baseline-combined
#| tbl-cap: "Baseline (pre-construction) comparability of stone pile, control, and reference sites. Values are mean ± SD; p-values test Stone pile vs Control (paired Wilcoxon signed-rank test) and Stone pile vs Reference (unpaired Wilcoxon rank-sum test)."

# Baseline-only data (monitoring round 0), all three site types
baseline_data <- all_data %>%
  filter(monitoring_int == 0) %>%
  mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

# Display order and nicer names for the paper 
var_labels <- c(
  Weighted_CPUE_Kōura    = "Kōura CPUE",
  Weighted_CPUE_Catfish  = "Catfish CPUE",
  Weighted_BPUE_Kōura    = "Kōura BPUE",
  Weighted_BPUE_Catfish  = "Catfish BPUE",
  Substrate_index_before = "Substrate index (before)",
  Substrate_index_after  = "Substrate index (after)",
  Slope_0_20m            = "Slope (0–20 m)",
  Temperature            = "Temperature (°C)",
  DO_mgl                 = "Dissolved oxygen (mg/L)",
  Specific_conductivity  = "Specific conductivity (µS/cm)",
  pH                     = "pH")

baseline_vars <- setdiff(names(var_labels), "Substrate_index_after")

# Stone pile vs Control — paired by pair_id (BACI design check).
paired_test <- function(var, data = baseline_data) {
  wide <- data %>%
    dplyr::filter(site %in% c("Stone_pile", "Control")) %>%
    dplyr::select(pair_id, site, value = dplyr::all_of(var)) %>%
    tidyr::pivot_wider(names_from = site, values_from = value)

  test <- suppressWarnings(wilcox.test(wide$Stone_pile, wide$Control, paired = TRUE, exact = FALSE))
  tibble::tibble(Comparison = "Stone pile vs Control (paired)", Variable = var,
                 n_pairs = nrow(wide), statistic = unname(test$statistic), p.value = test$p.value)
}

# Stone pile vs Reference — unpaired (documents the baseline gap for RQ2)
unpaired_test <- function(var, data = baseline_data) {
  d <- data %>%
    dplyr::filter(site %in% c("Stone_pile", "Reference")) %>%
    dplyr::select(site, value = dplyr::all_of(var)) %>%
    dplyr::filter(!is.na(value)) %>%
    dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Reference")))

  test <- suppressWarnings(wilcox.test(value ~ site, data = d, exact = FALSE))
  tibble::tibble(Comparison = "Stone pile vs Reference (unpaired)", Variable = var,
                 n_pairs = NA_integer_, statistic = unname(test$statistic), p.value = test$p.value)
}

baseline_tests <- dplyr::bind_rows(
  purrr::map_dfr(baseline_vars, paired_test),
  purrr::map_dfr(baseline_vars, unpaired_test)) %>%
  dplyr::mutate(across(c(statistic, p.value), ~ round(.x, 3)))

#write.csv(baseline_tests, file.path(out_dir, "tbl-baseline-tests.csv"), row.names = FALSE)

# Combined view
baseline_group_summary <- baseline_data %>%
  dplyr::select(site, dplyr::all_of(baseline_vars)) %>%
  tidyr::pivot_longer(-site, names_to = "Variable", values_to = "Value") %>%
  dplyr::group_by(Variable, site) %>%
  dplyr::summarise(
    mean_sd = paste0(round(mean(Value, na.rm = TRUE), 2), " ± ", round(sd(Value, na.rm = TRUE), 2)),
    .groups = "drop") %>%
  tidyr::pivot_wider(names_from = site, values_from = mean_sd)

baseline_tests_wide <- baseline_tests %>%
  dplyr::mutate(Comparison = dplyr::if_else(
    stringr::str_detect(Comparison, "Control"), "p_vs_Control", "p_vs_Reference")) %>%
  dplyr::select(Variable, Comparison, p.value) %>%
  tidyr::pivot_wider(names_from = Comparison, values_from = p.value)

baseline_combined <- baseline_group_summary %>%
  dplyr::left_join(baseline_tests_wide, by = "Variable") %>%
  dplyr::select(Variable, Stone_pile, Control, Reference, p_vs_Control, p_vs_Reference)

after_data <- all_data %>%
  filter(monitoring_int >= 1) %>%
  distinct(site_id, pair_id, site, Substrate_index_after) %>%
  mutate(site = factor(site, levels = c("Stone_pile", "Control", "Reference")))

after_tests <- dplyr::bind_rows(
  paired_test("Substrate_index_after", data = after_data),
  unpaired_test("Substrate_index_after", data = after_data)) %>%
  dplyr::mutate(across(c(statistic, p.value), ~ round(.x, 3)))

after_group_summary <- after_data %>%
  dplyr::select(site, Substrate_index_after) %>%
  tidyr::pivot_longer(-site, names_to = "Variable", values_to = "Value") %>%
  dplyr::group_by(Variable, site) %>%
  dplyr::summarise(
    mean_sd = paste0(round(mean(Value, na.rm = TRUE), 2), " ± ", round(sd(Value, na.rm = TRUE), 2)),
    .groups = "drop") %>%
  tidyr::pivot_wider(names_from = site, values_from = mean_sd)

after_tests_wide <- after_tests %>%
  dplyr::mutate(Comparison = dplyr::if_else(
    stringr::str_detect(Comparison, "Control"), "p_vs_Control", "p_vs_Reference")) %>%
  dplyr::select(Variable, Comparison, p.value) %>%
  tidyr::pivot_wider(names_from = Comparison, values_from = p.value)

after_combined <- after_group_summary %>%
  dplyr::left_join(after_tests_wide, by = "Variable") %>%
  dplyr::select(Variable, Stone_pile, Control, Reference, p_vs_Control, p_vs_Reference)

baseline_combined <- dplyr::bind_rows(baseline_combined, after_combined) %>%
  dplyr::mutate(Variable = factor(Variable, levels = names(var_labels))) %>%
  dplyr::arrange(Variable) %>%
  dplyr::mutate(Variable = var_labels[as.character(Variable)])

write.csv(baseline_combined, file.path(out_dir, "tbl-baseline-combined.csv"), row.names = FALSE)

knitr::kable(baseline_combined, digits = 3)
```


## Model constants
```{r}
#| label: model-constants
#| include: false

all_species <- c("Catfish", "Kōura")

species_letter_map  <- c("Catfish" = "c", "Kōura" = "k")
response_letter_map <- c("abundance" = "a", "biomass" = "b", "occupancy" = "o")

# Fix model$call so DHARMa/emmeans can re-find the named dataset
repoint_call <- function(m, form, data_name, family_expr) {
  if (is.null(m)) return(NULL)
  m$call$formula <- as.formula(form)
  m$call$data    <- as.name(data_name)
  m$call$family  <- family_expr
  m
}
```

## RQ1 — Baseline
```{r}
#| label: tbl-rq1-full
#| tbl-cap: "RQ1"

rq1a_data <- all_data %>%
  dplyr::filter(site %in% c("Stone_pile", "Control"), monitoring == 0) %>%
  dplyr::mutate(
    Site_type_detail = factor(Site_type_detail, levels = c("Control", "Stone_pile")))

rq1b_data <- all_data %>%
  dplyr::filter(site == "Reference", monitoring == 0) %>%
  dplyr::mutate(
    Site_type_detail = factor(Site_type_detail, levels = c("Reference_1", "Reference_2")))

fit_rq1 <- function(data, species, response_type) {
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species) else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(data)) return(NULL)
  form <- paste0("`", response_col, "` ~ Site_type_detail")
  tryCatch(
    glmmTMB::glmmTMB(as.formula(form), data = data, family = tweedie()),
    error = function(e) NULL)
}

fit_rq1b <- function(data, species, response_type) {
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species) else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(data)) return(NULL)
  # Temperature supported for reference catfish: M1bac Base fails; M1bbc ΔAICc=5
  form <- if (species == "Catfish")
    paste0("`", response_col, "` ~ Site_type_detail + Temperature")
  else
    paste0("`", response_col, "` ~ Site_type_detail")
  tryCatch(
    glmmTMB::glmmTMB(as.formula(form), data = data, family = tweedie()),
    error = function(e) NULL)
}

# Helper to build a results tibble for one sub-question 
make_rq1_results <- function(data, fit_fn, comparison, time_frame, term_of_interest, model_prefix) {
  tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
    dplyr::mutate(
      RQ               = "RQ1",
      comparison       = comparison,
      time_frame       = time_frame,
      term_of_interest = term_of_interest,
      model_id         = paste0(model_prefix, response_letter_map[response_type], species_letter_map[species]),
      model            = purrr::map2(species, response_type, ~ fit_fn(data, .x, .y))) %>%
    dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
    dplyr::mutate(
      p_value  = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
        if (is.null(m)) return(NA_real_)
        broom.mixed::tidy(m, effects = "fixed", component = "cond") %>%
          dplyr::filter(term == trm) %>%
          dplyr::pull(p.value)
      }),
      p_adj_BH = p.adjust(p_value, method = "BH"))
}

rq1a_results <- make_rq1_results(
  rq1a_data, fit_rq1,
  comparison = "treatment", time_frame = "before",
  term_of_interest = "Site_type_detailStone_pile",
  model_prefix = "M1a")

rq1b_results <- make_rq1_results(
  rq1b_data, fit_rq1b,
  comparison = "reference", time_frame = "before",
  term_of_interest = "Site_type_detailReference_2",
  model_prefix = "M1b")

rq1_results <- dplyr::bind_rows(rq1a_results, rq1b_results)

rq1_full_tidy <- rq1_results %>%
  dplyr::mutate(
    time_frame  = factor(time_frame,  levels = c("before", "after")),
    comparison  = factor(comparison,  levels = c("treatment", "reference")),
    species     = factor(species,     levels = names(species_letter_map)),
    response_type = factor(response_type, levels = names(response_letter_map))) %>%
  dplyr::arrange(time_frame, comparison, species, response_type) %>%
  dplyr::mutate(tidy = purrr::map(model, ~ if (is.null(.x)) NULL else
    broom.mixed::tidy(.x, effects = "fixed", component = "cond", conf.int = TRUE))) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, comparison, time_frame, species, response_type,
                term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

print(rq1_full_tidy, n=18)

DT::datatable(
  rq1_full_tidy,
  filter = "top",
  options = list(pageLength = 16, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)


```

## RQ2 — Construction impact
```{r}
#| label: tbl-rq2-full
#| tbl-cap: "RQ2"

# RQ2 a --------
rq2a_data <- all_data %>%
  filter(site %in% c("Stone_pile", "Control")) %>%
  mutate(
    Period  = factor(Period, levels = c("Before", "After")),
    site    = factor(site,   levels = c("Control", "Stone_pile")),
    site_id = factor(site_id))

# fit function for RQ2 models
fit_rq2a <- function(species, response_type) {
  response_col <- dplyr::case_when(
    response_type == "occupancy" ~ paste0("Presence_", species),
    response_type == "abundance" ~ paste0("Weighted_CPUE_", species),
    response_type == "biomass"   ~ paste0("Weighted_BPUE_", species))
  if (!response_col %in% names(rq2a_data)) return(NULL)

  # Season collinear with Period in BACI; catfish models include fetch_z (AICc-selected)
  fetch_term <- if (species == "Catfish") " + fetch_z" else ""
  form <- paste0("`", response_col, "` ~ Period * site + monitoring_int", fetch_term, " + (1 | pair_id / site_id)")
  fam_obj <- if (response_type == "occupancy") binomial() else tweedie()
  zif     <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = rq2a_data, REML = TRUE, family = fam_obj,
                  ziformula = zif),
                error = function(e) NULL)
  if (!is.null(m) && m$fit$convergence != 0)
    m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = rq2a_data, REML = TRUE, family = fam_obj,
                    ziformula = zif,
                    control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
                  error = function(e) NULL)
  repoint_call(m, form, "rq2a_data",
               if (response_type == "occupancy") quote(binomial()) else quote(tweedie()))
}

rq2a_results <- tidyr::crossing(species = all_species, response_type = c("occupancy", "abundance", "biomass")) %>%
  dplyr::mutate(
    RQ                = "RQ2a",
    comparison        = "vs_control",
    time_frame        = "full",
    term_of_interest  = "PeriodAfter:siteStone_pile",
    model_id          = paste0("M2a", response_letter_map[response_type], species_letter_map[species]),
    model             = purrr::map2(species, response_type, fit_rq2a)) %>%

  dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
  dplyr::mutate(
    p_value = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
      if (is.null(m)) return(NA_real_)
      broom.mixed::tidy(m, effects = "fixed", component = "cond") %>%
        dplyr::filter(term == trm) %>%
        dplyr::pull(p.value)
    }),
    p_adj_BH = p.adjust(p_value, method = "BH"))


# RQ2b ----------
rq2b_data <- rq2a_data %>%
  dplyr::filter(monitoring_int >= 1)

# fit function for RQ2b models
fit_rq2b <- function(species, response_type) {
  response_col <- dplyr::case_when(
    response_type == "occupancy" ~ paste0("Presence_", species),
    response_type == "abundance" ~ paste0("Weighted_CPUE_", species),
    response_type == "biomass"   ~ paste0("Weighted_BPUE_", species))
  if (!response_col %in% names(rq2b_data)) return(NULL)

  fam_obj <- if (response_type == "occupancy") binomial() else tweedie()
  zif     <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  # Season collinear with Period in BACI context; catfish models include fetch_z (AICc-selected)
  fetch_term <- if (species == "Catfish") " + fetch_z" else ""
  form <- paste0("`", response_col, "` ~ site * monitoring_int", fetch_term, " + (1 | pair_id/site_id)")
  m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = rq2b_data, REML = TRUE, family = fam_obj,
                  ziformula = zif),
                error = function(e) NULL)
  if (!is.null(m) && m$fit$convergence != 0)
    m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = rq2b_data, REML = TRUE, family = fam_obj,
                    ziformula = zif,
                    control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
                  error = function(e) NULL)
  repoint_call(m, form, "rq2b_data",
               if (response_type == "occupancy") quote(binomial()) else quote(tweedie()))
}

rq2b_results <- tidyr::crossing(species = all_species, response_type = c("occupancy", "abundance", "biomass")) %>%
  dplyr::mutate(
    RQ                = "RQ2b",
    comparison        = "vs_control",
    time_frame        = "after",
    term_of_interest  = "siteStone_pile:monitoring_int",
    model_id          = paste0("M2b", response_letter_map[response_type], species_letter_map[species]),
    model             = purrr::map2(species, response_type, fit_rq2b)) %>%

  dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
  dplyr::mutate(
    p_value = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
      if (is.null(m)) return(NA_real_)
      broom.mixed::tidy(m, effects = "fixed", component = "cond") %>%
        dplyr::filter(term == trm) %>%
        dplyr::pull(p.value)
    }),
    p_adj_BH = p.adjust(p_value, method = "BH"))

rq2_results <- dplyr::bind_rows(rq2a_results, rq2b_results)

rq2_full_tidy <- rq2_results %>%
  dplyr::mutate(tidy = purrr::map(model, ~ if (is.null(.x)) NULL else broom.mixed::tidy(.x, effects = "fixed", component = "cond", conf.int = TRUE))) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, species, response_type, term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

print(rq2_full_tidy, n=70)

DT::datatable(
  rq2_full_tidy,
  filter = "top",
  options = list(pageLength = 70, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)



```

```{r}

```


## RQ3 — Recovery toward reference conditions
```{r}
#| label: tbl-rq3-full
#| tbl-cap: "RQ3"

# RQ3a ----------
rq3a_data <- all_data %>%
  dplyr::filter(site %in% c("Stone_pile", "Reference"), monitoring_int >= 1) %>%
  dplyr::mutate(
    site    = factor(site,   levels = c("Reference", "Stone_pile")),
    site_id = factor(site_id))

fit_rq3a <- function(species, response_type) {
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species) else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(rq3a_data)) return(NULL)

  zif  <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  # AICc selection: Temperature strongly supported for catfish abundance (ΔAICc=14.4); all others → Base
  form <- if (species == "Catfish" && response_type == "abundance")
    paste0("`", response_col, "` ~ site * monitoring_int + Temperature + (1 | site_id)")
  else
    paste0("`", response_col, "` ~ site * monitoring_int + (1 | site_id)")
  m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = rq3a_data, REML = TRUE, family = tweedie(),
                  ziformula = zif),
                error = function(e) NULL)
  if (!is.null(m) && m$fit$convergence != 0)
    m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = rq3a_data, REML = TRUE, family = tweedie(),
                    ziformula = zif,
                    control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
                  error = function(e) NULL)
  repoint_call(m, form, "rq3a_data", quote(tweedie()))
}

rq3a_results <- tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
  dplyr::mutate(
    RQ                = "RQ3a",
    comparison        = "vs_reference",
    time_frame        = "after",
    term_of_interest  = "siteStone_pile:monitoring_int",
    model_id          = paste0("M3a", response_letter_map[response_type], species_letter_map[species]),
    model             = purrr::map2(species, response_type, fit_rq3a)) %>%
  dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
  dplyr::mutate(
    p_value = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
      if (is.null(m)) return(NA_real_)
      broom.mixed::tidy(m, effects = "fixed", component = "cond") %>%
        dplyr::filter(term == trm) %>%
        dplyr::pull(p.value)
    }),
    p_adj_BH = p.adjust(p_value, method = "BH"))

# RQ3b ----------
rq3b_data <- all_data %>%
  dplyr::filter(site == "Reference", monitoring_int >= 1) %>%
  dplyr::mutate(
    Site_type_detail = factor(Site_type_detail, levels = c("Reference_1", "Reference_2")),
    pair_id = factor(pair_id), site_id = factor(site_id))

fit_rq3b_inner <- function(data, species, response_type) {
  response_col <- if (response_type == "abundance") paste0("Weighted_CPUE_", species) else paste0("Weighted_BPUE_", species)
  if (!response_col %in% names(data)) return(NULL)
  zif  <- if (species == "Catfish" && response_type == "abundance") ~1 else ~0
  # AICc selection: Season for kōura abundance (ΔAICc=4.7); Temperature for catfish abundance; Base for biomass
  form <- if (species == "Kōura" && response_type == "abundance")
    paste0("`", response_col, "` ~ Site_type_detail * monitoring_int + Season + (1 | pair_id / site_id)")
  else if (response_type == "abundance")
    paste0("`", response_col, "` ~ Site_type_detail * monitoring_int + Temperature + (1 | pair_id / site_id)")
  else
    paste0("`", response_col, "` ~ Site_type_detail * monitoring_int + (1 | pair_id / site_id)")

  conv_failed <- FALSE
  m <- withCallingHandlers(
    tryCatch(
      glmmTMB::glmmTMB(as.formula(form), data = data, REML = TRUE, family = tweedie(),
        ziformula = zif),
      error = function(e) NULL),
    warning = function(w) {
      if (grepl("convergence|false conv", conditionMessage(w), ignore.case = TRUE))
        conv_failed <<- TRUE
      invokeRestart("muffleWarning")
    })

  if (conv_failed || is.null(m)) {
    m <- tryCatch(
      glmmTMB::glmmTMB(as.formula(form), data = data, REML = TRUE, family = tweedie(),
        ziformula = zif,
        control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
      error = function(e) NULL)
  }
  m
}

# Helper to build a results tibble for one sub-question
make_rq3b_results <- function(data, fit_fn, comparison, time_frame, term_of_interest, model_prefix) {
  tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
    dplyr::mutate(
      RQ               = "RQ3b",
      comparison       = comparison,
      time_frame       = time_frame,
      term_of_interest = term_of_interest,
      model_id         = paste0(model_prefix, response_letter_map[response_type], species_letter_map[species]),
      model            = purrr::map2(species, response_type, ~ fit_fn(data, .x, .y))) %>%
    dplyr::arrange(match(species, names(species_letter_map)), match(response_type, names(response_letter_map))) %>%
    dplyr::mutate(
      p_value  = purrr::map2_dbl(model, term_of_interest, function(m, trm) {
        if (is.null(m)) return(NA_real_)
        broom.mixed::tidy(m, effects = "fixed", component = "cond") %>%
          dplyr::filter(term == trm) %>%
          dplyr::pull(p.value)
      }),
      p_adj_BH = p.adjust(p_value, method = "BH"))
}

rq3b_results <- make_rq3b_results(
  rq3b_data, fit_rq3b_inner,
  comparison = "reference", time_frame = "after",
  term_of_interest = "Site_type_detailReference_2:monitoring_int",
  model_prefix = "M3b")

rq3_results <- dplyr::bind_rows(rq3a_results, rq3b_results)

rq3_full_tidy <- rq3_results %>%
  dplyr::mutate(tidy = purrr::map(model, ~ if (is.null(.x)) NULL else broom.mixed::tidy(.x, effects = "fixed", component = "cond", conf.int = TRUE))) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, species, response_type, term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

print(rq3_full_tidy, n=66)

DT::datatable(
  rq3_full_tidy,
  filter = "top",
  options = list(pageLength = 34, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)



```


### Predicted trajectories (kōura abundance convergence)
```{r}
#| label: fig-convergence-plot
#| fig-width: 8
#| fig-height: 4
#| dpi: 300
#| fig-cap: "Predicted kōura CPUE trajectories over monitoring rounds at stone pile and reference sites from the abundance convergence model. Shaded ribbons represent 95% confidence intervals. The annotation shows the key interaction term (β ± p-value) and model R² values for fixed effects (R²m) and conditional (R²c)."


M3aak <- rq3a_results %>%
  dplyr::filter(species == "Kōura", response_type == "abundance") %>%
  dplyr::pull(model) %>% .[[1]]

# Predicted trajectories at each monitoring round (M3aak: response ~ site * monitoring_int + (1|site_id))
emmeans::emm_options(msg.interaction = FALSE)

emm_m3aak <- purrr::map_dfr(1:6, function(i) {
  emmeans(M3aak, ~ site, at = list(monitoring_int = i), type = "response") %>%
    as.data.frame() %>%
    dplyr::mutate(monitoring_int = i)
}) %>%
  dplyr::rename(lower.CL = dplyr::any_of("asymp.LCL"),
                upper.CL = dplyr::any_of("asymp.UCL")) %>%
  mutate(site = factor(site, levels = c("Stone_pile", "Reference")))

# Clean numeric values for annotation
beta_val <- round(glmmTMB::fixef(M3aak)$cond["siteStone_pile:monitoring_int"], 3)
p_val    <- round(summary(M3aak)$coefficients$cond["siteStone_pile:monitoring_int", "Pr(>|z|)"], 3)
r2_vals  <- tryCatch(
  MuMIn::r.squaredGLMM(M3aak),
  error = function(e) matrix(c(NA_real_, NA_real_), nrow = 1,
                              dimnames = list(NULL, c("R2m", "R2c")))
)
r2m_val  <- round(r2_vals[1, "R2m"], 2)
r2c_val  <- round(r2_vals[1, "R2c"], 2)

# Use plotmath-safe expression
model_label <- paste0("italic(beta)[Stone_pile%*%time]==", beta_val, "~(italic(p)==", p_val, ")", "~~italic(R)[m]^2==", r2m_val, "~~italic(R)[c]^2==", r2c_val)

convergence_data_plot <- rq3a_data %>%
  filter(!is.na(`Weighted_CPUE_Kōura`)) %>%
  mutate(site = factor(site, levels = c("Stone_pile", "Reference")))

pre_data_plot <- all_data %>%
  dplyr::filter(site %in% c("Stone_pile", "Reference"),
                monitoring_int == 0,
                !is.na(`Weighted_CPUE_Kōura`)) %>%
  dplyr::mutate(site = factor(site, levels = c("Stone_pile", "Reference")))

convergence_plot <- ggplot() +
  geom_vline(xintercept = 0.25, linetype = "dashed", colour = "grey40", linewidth = 0.6) +
  annotate("text", x = -0.15, y = Inf, label = "Before", vjust = 1.5,
           hjust = 0.5, size = 3, colour = "grey40") +
  annotate("text", x = 3.5, y = Inf, label = "After", vjust = 1.5,
           hjust = 0.5, size = 3, colour = "grey40") +
  geom_point(data = pre_data_plot, aes(monitoring_int, `Weighted_CPUE_Kōura`, fill = site, shape = site),
             colour = "black", alpha = 0.2, size = 1.8, stroke = 0.5,
             position = position_jitter(width = 0.08, seed = 42), show.legend = FALSE) +
  geom_point(data = convergence_data_plot, aes(monitoring_int, `Weighted_CPUE_Kōura`,  fill = site, shape = site),  colour = "black",  alpha = 0.4, size = 1.8, stroke = 0.5, position = position_jitter(width = 0.08, seed = 42),show.legend = FALSE) +
  geom_ribbon(data = emm_m3aak, aes(monitoring_int, response,  ymin = lower.CL, ymax = upper.CL, fill = site),  alpha = 0.15, colour = NA,  show.legend = FALSE) +
  geom_line(data = emm_m3aak,  aes(monitoring_int, response,  colour = site, group = site), linewidth = 1.1) +
  geom_point(data = emm_m3aak, aes(monitoring_int, response,  fill = site, shape = site), colour = "black", size = 3, stroke = 0.4) +
  annotate("text", x = 0.5, y = Inf, label = model_label, parse = TRUE,  hjust = 0, vjust = 1.8, size = 3.2, colour = "grey30") +
  scale_colour_manual(values = conv_colours, labels = conv_labels) +
  scale_fill_manual(  values = conv_fill,    labels = conv_labels) +
  scale_shape_manual( values = conv_shapes,  labels = conv_labels) +
  scale_x_continuous(breaks = 0:6, labels = monitor_labels, limits = c(-0.3, 6.3)) +
  scale_y_continuous(trans = "log1p") +
  labs(x=NULL ,y= "Kōura CPUE (log scale)",colour = "Site type",fill= "Site type",shape  = "Site type") +
  theme(axis.text.x = element_text(size = 9), axis.text.y = element_text(size = 9), axis.title.y = element_text(size = 10), legend.position = "right")

ggsave(file.path(out_dir, "fig-convergence-plot.png"), convergence_plot, width = 6, height = 3, dpi = 300)

convergence_plot
```


### Before after Model figure
```{r}
#| label: fig-before-after-models
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "Observed distributions of kōura and catfish CPUE and BPUE by site type and monitoring period (before/after stone pile installation). Boxes show interquartile range, line = median, whiskers extend to 1.5×IQR."

before_after_combined <- dplyr::bind_rows(
  dplyr::mutate(rq1a_data,  period = "Before"),
  dplyr::mutate(rq2b_data,  period = "After"),
  dplyr::mutate(rq1b_data,  period = "Before"),
  dplyr::mutate(rq3b_data,  period = "After")) |>
  dplyr::mutate(
    period          = factor(period, levels = c("Before", "After")),
    Site_type_detail = factor(Site_type_detail,
      levels = c("Stone_pile", "Control", "Reference_1", "Reference_2")))

response_cols <- intersect(
  c(paste0("Weighted_CPUE_", all_species), paste0("Weighted_BPUE_", all_species)),
  names(before_after_combined))

before_after_long <- before_after_combined |>
  tidyr::pivot_longer(cols = dplyr::all_of(response_cols),
                      names_to = "response_var", values_to = "value") |>
  dplyr::mutate(
    species  = stringr::str_extract(response_var,
                 paste(all_species, collapse = "|")),
    response = dplyr::if_else(stringr::str_detect(response_var, "CPUE"),
                              "Abundance (CPUE)", "Biomass (BPUE)"),
    panel    = factor(paste0(species, " — ", response),
                levels = c(paste0(all_species[1], " — Abundance (CPUE)"),
                           paste0(all_species[1], " — Biomass (BPUE)"),
                           paste0(all_species[2], " — Abundance (CPUE)"),
                           paste0(all_species[2], " — Biomass (BPUE)"))))

ggplot2::ggplot(before_after_long,
    ggplot2::aes(x = Site_type_detail, y = value, fill = period)) +
  ggplot2::geom_boxplot(outlier.size = 0.6,
                        position = ggplot2::position_dodge(0.8)) +
  ggplot2::facet_wrap(~ panel, scales = "free_y", ncol = 2) +
  ggplot2::scale_fill_manual(values = c("Before" = "#c9a96e", "After" = "#3d7a5e")) +
  ggplot2::scale_x_discrete(labels = c(
    "Stone_pile"  = "Stone pile",
    "Control"     = "Control",
    "Reference_1" = "Reference 1",
    "Reference_2" = "Reference 2")) +
  ggplot2::scale_y_continuous(trans = "log1p") +
  ggplot2::labs(x = NULL, y = "log(value + 1)", fill = "Period") +
  ggplot2::theme_bw() +
  ggplot2::theme(
    strip.text      = ggplot2::element_text(face = "bold"),
    axis.text.x     = ggplot2::element_text(angle = 35, hjust = 1),
    legend.position = "bottom")
```

## Model diagnostics 
```{r}
#| label: fig-residual-diagnostics
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "DHARMa scaled-residual diagnostics for each model: QQ plot (left) and residual vs. predicted (right). Red flags indicate a significant departure from the expected uniform distribution, dispersion, or outlier rate."

model_registry <- dplyr::bind_rows(rq1_results, rq2_results, rq3_results) %>%
  dplyr::select(model_id, RQ, species, response_type, comparison, time_frame, term_of_interest, p_adj_BH, model)


is_dharma_safe <- function(m) {
  if (is.null(m) || m$fit$convergence != 0) return(FALSE)
  tryCatch({
    v <- diag(as.matrix(vcov(m, full = TRUE)))
    all(is.finite(v) & v > 0)
  }, error = function(e) FALSE, warning = function(w) FALSE)
}

diagnostic_models <- model_registry %>%
  dplyr::filter(time_frame != "before") %>%
  dplyr::filter(purrr::map_lgl(model, is_dharma_safe)) %>%
  { purrr::set_names(.$model, .$model_id) } %>%
  purrr::compact()

simulate_check <- function(model, name, seed = 42) {
  sim <- DHARMa::simulateResiduals(model, n = 1000, seed = seed)
  plot(sim, main = name)
  sim
}

dharma_sims <- purrr::imap(diagnostic_models, simulate_check)
```

### Residual diagnostics summary table
```{r}
#| label: tbl-residual-diagnostics
#| tbl-cap: "DHARMa goodness-of-fit tests per model. p < 0.05 indicates a significant departure from model assumptions (uniformity of residuals, over/under-dispersion, or excess outliers) and warrants closer inspection of the corresponding diagnostic plot above."

dharma_tests <- purrr::imap_dfr(dharma_sims, function(sim, name) {
  tibble(
    Model           = name,
    Uniformity_p    = DHARMa::testUniformity(sim, plot = FALSE)$p.value,
    Dispersion_p    = DHARMa::testDispersion(sim, plot = FALSE)$p.value,
    Outlier_p       = DHARMa::testOutliers(sim, plot = FALSE)$p.value)
}) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(dharma_tests)

write.csv(dharma_tests, file.path(out_dir, "tbl-residual-diagnostics.csv"), row.names = FALSE)
```

# Model results
## Question structure reference table
```{r}
#| label: tbl-question-structure
#| tbl-cap: "Model structure for each research question, species, and response type. RQ4 after-period formulas vary by species because Temperature was retained or dropped based on AICc model selection and Hessian validity per model. All models fit via glmmTMB (Tweedie for abundance/biomass, binomial for occupancy)."

question_structure <- model_registry %>%
  dplyr::mutate(
    family = dplyr::if_else(response_type == "occupancy", "binomial", "tweedie"),
    data = dplyr::case_when(
      comparison == "vs_control"  & time_frame == "full"   ~ "rq2a_data",
      comparison == "vs_control"  & time_frame == "after"  ~ "rq2b_data",
      comparison == "vs_reference"                         ~ "rq3a_data",
      comparison == "treatment"   & time_frame == "before" ~ "rq1a_data",
      comparison == "reference"   & time_frame == "before" ~ "rq1b_data",
      comparison == "reference"   & time_frame == "after"  ~ "rq3b_data"),
    formula = purrr::map_chr(model, function(m) {
      if (is.null(m)) return(NA_character_)
      f <- tryCatch(deparse(formula(m)), error = function(e) NA_character_)
      stringr::str_replace(paste(f, collapse = " "), "^`[^`]+`", "<response>")
    })) %>%
  dplyr::mutate(
    time_frame = factor(time_frame, levels = c("full", "before", "after")),
    comparison = factor(comparison, levels = c("vs_control", "vs_reference", "treatment", "reference"))) %>%
  dplyr::arrange(RQ, time_frame, comparison, species, response_type) %>%
  dplyr::select(RQ, species, response_type, comparison, time_frame, family, data, formula, term_of_interest)

write.csv(question_structure, file.path(out_dir, "tbl-question-structure.csv"), row.names = FALSE)

knitr::kable(question_structure)

```


## Models summary table
```{r}
#| label: tbl-models-summary
#| include: true
#| tbl-cap: "Key interaction-term estimate (β, SE, 95% CI, BH-adjusted p-value) for every model in model_registry (RQ1–RQ4; occupancy/abundance/biomass; kōura and catfish). Models with non-positive-definite Hessian are included but will have NA confidence intervals."

tidy_all <- model_registry %>%
  dplyr::mutate(tidy = purrr::map2(model, term_of_interest, function(m, trm) {
    if (is.null(m)) return(NULL)
    broom.mixed::tidy(m, effects = "fixed", component = "cond", conf.int = TRUE) %>%
      dplyr::filter(term == trm)
  })) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, RQ, species, response_type, comparison, time_frame,
                term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

print(tidy_all, n=30)

DT::datatable(
  tidy_all,
  filter = "top",
  options = list(pageLength = 30, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)

```

## Forest plot companion table
```{r}
#| label: tbl-forest-summary
#| tbl-cap: "Summary of key model results shown in the forest plot. Each row corresponds to one panel section and species. β is the estimate for the key term of interest (± 95% CI); direction indicates the sign of the effect regardless of significance. Significance is based on BH-adjusted p-values across all models."

section_labels <- c(
  "RQ1_treatment_before"   = "Before: stone pile vs control",
  "RQ1_reference_before"   = "Before: reference 1 vs reference 2",
  "RQ2_vs_control_full"    = "Before–After: BACI",
  "RQ2_vs_control_after"   = "After: trajectory",
  "RQ3_vs_reference_after" = "Convergence toward reference",
  "RQ3_reference_after"    = "Reference stability")

question_text <- c(
  "Before: stone pile vs control"      = "Were stone pile and control sites comparable before construction?",
  "Before: reference 1 vs reference 2" = "Were reference groups comparable at baseline?",
  "Before–After: BACI"                 = "Was the before-to-after change greater at stone pile than control sites?",
  "After: trajectory"                  = "Did stone pile and control sites diverge over the post-construction period?",
  "Convergence toward reference"       = "Did stone pile sites converge toward reference site levels over time?",
  "Reference stability"                = "Were reference site conditions stable across monitoring rounds?")

forest_summary <- tidy_all %>%
  dplyr::mutate(
    section_key = paste(sub("[ab]$", "", RQ), comparison, time_frame, sep = "_"),
    Section     = dplyr::recode(section_key, !!!section_labels),
    Question    = dplyr::recode(Section, !!!question_text),
    Result = dplyr::case_when(
      p_adj_BH < 0.05 & estimate > 0 ~ "Significant positive",
      p_adj_BH < 0.05 & estimate < 0 ~ "Significant negative",
      estimate > 0                    ~ "Non-significant positive",
      estimate < 0                    ~ "Non-significant negative",
      TRUE                            ~ "—"),
    CI = paste0(round(estimate, 2), " [", round(conf.low, 2), ", ", round(conf.high, 2), "]")) %>%
  dplyr::filter(Section %in% names(question_text)) %>%
  dplyr::mutate(
    Section = factor(Section, levels = names(question_text)),
    species  = factor(species, levels = c("Kōura", "Catfish")),
    response_type = factor(response_type, levels = c("abundance", "biomass", "occupancy"))) %>%
  dplyr::arrange(Section, species, response_type) %>%
  dplyr::select(Section, Question, species, response_type, CI, Result) %>%
  dplyr::rename(Species = species, Response = response_type,
                `β [95% CI]` = CI)

write.csv(forest_summary, file.path(out_dir, "tbl-forest-summary.csv"), row.names = FALSE)
knitr::kable(forest_summary, align = c("l","l","l","l","r","l"))
```

## Full fixed-effects table
```{r}
#| label: tbl-models-summary-full
#| include: true
#| tbl-cap: "Full fixed-effect estimates (β, SE, 95% CI, p-value) for every term in every model."

tidy_all_full <- model_registry %>%
  dplyr::mutate(tidy = purrr::map(model, function(m) {
    if (is.null(m)) return(NULL)
    broom.mixed::tidy(m, effects = "fixed", component = "cond", conf.int = TRUE)
  })) %>%
  dplyr::filter(!purrr::map_lgl(tidy, is.null)) %>%
  tidyr::unnest(tidy) %>%
  # p_adj_BH only applies to each model's own term of interest — NA elsewhere.
  dplyr::mutate(p_adj_BH = dplyr::if_else(term == term_of_interest, p_adj_BH, NA_real_)) %>%
  dplyr::select(model_id, RQ, species, response_type, comparison, time_frame,
                term, estimate, std.error, statistic, p.value, p_adj_BH, conf.low, conf.high) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

write.csv(tidy_all_full, file.path(out_dir, "tbl-models-summary-full.csv"), row.names = FALSE)

# Save M2aak variance components for occupancy_analysis.qmd
m2aak_obj <- rq2a_results$model[[which(rq2a_results$model_id == "M2aak")]]
if (!is.null(m2aak_obj)) {
  vc_df <- as.data.frame(VarCorr(m2aak_obj)$cond)
  tibble::tibble(
    sigma_pair  = vc_df$sdcor[vc_df$grp == "pair_id"],
    sigma_site  = vc_df$sdcor[vc_df$grp == "site_id:pair_id"],
    sigma_resid = sigma(m2aak_obj)
  ) |>
    readr::write_csv(file.path(der_data_dir, "M2aak_variance_components.csv"))
}

print(tidy_all_full, n=226)

DT::datatable(
  tidy_all_full,
  filter = "top",
  options = list(pageLength = 51, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons',
  rownames = FALSE
) |>
  DT::formatRound(columns = c("estimate","std.error","statistic","p.value","p_adj_BH","conf.low","conf.high"), digits = 3)
```

## All models forest plot
```{r}
#| label: fig-forest-plot
#| fig-width: 6
#| fig-height: 8
#| dpi: 300
#| fig-cap: "Effect sizes (β ± 95% CI) for the key term of interest across all models, separated by species (columns) and comparison type (rows). Filled points indicate significant effects (*p* < 0.05); open points are non-significant. Each row has its own x-axis scale — effect magnitudes are not directly comparable across rows because each comparison tests a different contrast (step-change, per-round slope, or simple group difference), and occupancy is on the log-odds scale while abundance and biomass models use the Tweedie log link. Models with degenerate confidence intervals are excluded."

group_levels <- c(
  "Were stone pile and control sites\ncomparable before construction?",
  "Were the two reference groups\ncomparable at baseline?",
  "Was the before-after change greater at\nstone pile than control sites?",
  "Did stone pile and control sites diverge\nover the post-construction period?",
  "Did stone pile sites converge toward\nreference site densities over time?",
  "Were reference site conditions stable\nacross monitoring rounds?")

strip_fills <- setNames(
  c("#dce8f0", "#dce8f0", "#f8dfbf", "#f8dfbf", "#d6eddc", "#d6eddc"),
  group_levels)

forest_data <- tidy_all %>%
  dplyr::filter(is.finite(conf.low) & is.finite(conf.high)) %>%
  dplyr::mutate(
    row_label = dplyr::recode(response_type,
      "abundance" = "Abundance", "biomass" = "Biomass", "occupancy" = "Occupancy"),
    group = dplyr::case_when(
      RQ == "RQ1"  & comparison %in% c("vs_control", "treatment") ~ group_levels[1],
      RQ == "RQ1"  & comparison == "reference"                    ~ group_levels[2],
      RQ == "RQ2a"                                                ~ group_levels[3],
      RQ == "RQ2b"                                                ~ group_levels[4],
      RQ == "RQ3a"                                                ~ group_levels[5],
      RQ == "RQ3b"                                                ~ group_levels[6],
      TRUE ~ RQ),
    significant = p.value < 0.05) %>%
  dplyr::mutate(
    row_label = factor(row_label, levels = rev(c("Abundance", "Biomass", "Occupancy"))),
    group     = factor(group,     levels = group_levels),
    species   = factor(species,   levels = c("Kōura", "Catfish")))

make_species_forest <- function(sp_data) {
  grp_order <- levels(sp_data$group)
  fills     <- strip_fills[grp_order]
  ggplot(sp_data, aes(estimate, row_label)) +
    geom_vline(xintercept = 0, linetype = "dashed",
               colour = "grey40", linewidth = 0.5) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.25, linewidth = 0.6) +
    geom_point(data = dplyr::filter(sp_data,  significant),
               shape = 16, size = 3) +
    geom_point(data = dplyr::filter(sp_data, !significant),
               shape = 21, fill = "white", size = 3, stroke = 0.8) +
    ggh4x::facet_wrap2(~ group, ncol = 1, scales = "free",
      strip = ggh4x::strip_themed(
        background_x = ggh4x::elem_list_rect(fill = fills,
                                             colour = "black",
                                             linewidth = 0.4))) +
    labs(x = "Effect size (β ± 95% CI)", y = NULL) +
    theme(
      strip.text         = element_text(size = 8, face = "bold"),
      axis.text.y        = element_text(size = 9),
      axis.text.x        = element_text(size = 8),
      legend.position    = "none",
      panel.grid.major.x = element_line(colour = "grey92", linewidth = 0.3),
      panel.spacing.y    = unit(0.5, "lines"))
}

p_koura <- make_species_forest(dplyr::filter(forest_data, species == "Kōura")) +
  labs(title = "Kōura") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5))

p_catfish <- make_species_forest(dplyr::filter(forest_data, species == "Catfish")) +
  labs(title = "Catfish") +
  theme(plot.title  = element_text(size = 11, face = "bold", hjust = 0.5),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank())

forest_plot <- p_koura | p_catfish

ggsave(file.path(out_dir, "fig-forest-plot.png"), forest_plot, width = 7, height = 9, dpi = 300)

forest_plot
```

# Additional and sensitivity models
## Check various additional models
```{r}
#| label: additional-models

# test the effect of koura length.
m_size <- lmerTest::lmer(
  Mean_Length_Kōura ~  site * monitoring_int + (1 | site_id),
  data = all_data %>% filter(!is.na(Mean_Length_Kōura)))

# Data-quality sensitivity check
quality_sensitivity_check <- function(results, data_obj) {
  purrr::pmap_dfr(results, function(model_id, term_of_interest, model, ...) {
    if (is.null(model)) return(NULL)
    dat_sens <- data_obj %>% dplyr::filter(monitoring_quality == "complete")
    m_sens <- tryCatch(update(model, data = dat_sens), error = function(e) NULL)
    orig <- broom.mixed::tidy(model, effects = "fixed") %>%
      dplyr::filter(term == term_of_interest) %>%
      dplyr::mutate(model_id = model_id, version = "All data")
    if (is.null(m_sens)) return(orig)
    sens <- broom.mixed::tidy(m_sens, effects = "fixed") %>%
      dplyr::filter(term == term_of_interest)
    if (nrow(sens) == 0 || is.na(sens$std.error) || sens$std.error <= 0) return(orig)
    sens <- sens %>% dplyr::mutate(model_id = model_id, version = "Complete quality only")
    dplyr::bind_rows(orig, sens)
  })
}

quality_sensitivity_all <- dplyr::bind_rows(
  quality_sensitivity_check(rq2a_results, rq2a_data),
  quality_sensitivity_check(rq2b_results, rq2b_data),
  quality_sensitivity_check(rq3a_results, rq3a_data),
  quality_sensitivity_check(rq3b_results, rq3b_data)) %>%
  dplyr::select(model_id, version, term, estimate, std.error, statistic, p.value) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(quality_sensitivity_all)

# Matched-season check, generalized to all 6 RQ2a models: baseline (monitoring 0, Pre/Summer 2025) vs. only the following Summer 2026 round (monitoring 4), instead of pooling all 5 post-construction rounds into "After". Simplified fixed-effects-only formula: with only 2 time points (n = 20), the full RQ2a formula (random effects + Temperature) fails to converge (non-positive-definite Hessian / singular fit).
baci_data_summer <- rq2a_data %>%
  dplyr::filter(monitoring_int %in% c(0, 4)) %>%
  dplyr::mutate(Period = factor(Period, levels = c("Before", "After")))

matched_season_check <- function(species, response_type) {
  response_col <- dplyr::case_when(
    response_type == "occupancy" ~ paste0("Presence_", species),
    response_type == "abundance" ~ paste0("Weighted_CPUE_", species),
    response_type == "biomass"   ~ paste0("Weighted_BPUE_", species))
  if (!response_col %in% names(baci_data_summer)) return(NULL)
  fam_obj <- if (response_type == "occupancy") binomial() else tweedie()
  form <- paste0("`", response_col, "` ~ Period * site")
  m <- tryCatch(glmmTMB::glmmTMB(as.formula(form), data = baci_data_summer, family = fam_obj),
                error = function(e) NULL)
  if (is.null(m)) return(NULL)
  broom.mixed::tidy(m, effects = "fixed") %>%
    dplyr::filter(term == "PeriodAfter:siteStone_pile") %>%
    dplyr::mutate(species = species, response_type = response_type)
}

matched_season_all <- purrr::map2_dfr(rq2a_results$species, rq2a_results$response_type, matched_season_check) %>%
  dplyr::select(species, response_type, term, estimate, std.error, statistic, p.value) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(matched_season_all)

# Substrate necessity check, generalized to every RQ2/RQ3 model: what
# happens to the term of interest if the substrate covariate is dropped
# entirely? RQ1 has no substrate term, so it's excluded here.
substrate_drop_check <- function(results, drop_term) {
  purrr::pmap_dfr(results, function(model_id, term_of_interest, model, ...) {
    if (is.null(model)) return(NULL)
    m_drop <- tryCatch(update(model, as.formula(paste(". ~ . -", drop_term))), error = function(e) NULL)
    orig <- broom.mixed::tidy(model, effects = "fixed") %>%
      dplyr::filter(term %in% c(term_of_interest, drop_term)) %>%
      dplyr::mutate(model_id = model_id, version = paste("With", drop_term))
    if (is.null(m_drop)) return(orig)
    dropped <- broom.mixed::tidy(m_drop, effects = "fixed") %>%
      dplyr::filter(term == term_of_interest) %>%
      dplyr::mutate(model_id = model_id, version = paste("Without", drop_term))
    dplyr::bind_rows(orig, dropped)
  })
}

substrate_necessity_all <- dplyr::bind_rows(
  substrate_drop_check(rq2_results, "Substrate_index_before"),
  substrate_drop_check(rq3_results, "Substrate_index_after")) %>%
  dplyr::select(model_id, version, term, estimate, std.error, statistic, p.value) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(substrate_necessity_all)

# Validate Season vs Temperature, generalized to all 4 RQ3 models:
# Temperature (a single linear predictor) previously failed DHARMa's
# residual-quantile test because kōura/catfish abundance cycles
# non-monotonically over the year — Season (categorical) can capture
# that, Temperature can't. Direct AICc comparison, swapping one for the
# other in the same base model. Only relevant to RQ3 — RQ1/RQ2 already
# use Temperature, not Season.
season_vs_temperature <- purrr::pmap_dfr(rq3_results, function(model_id, model, ...) {
  if (is.null(model)) return(NULL)
  m_temp <- tryCatch(update(model, . ~ . - Season + Temperature), error = function(e) NULL)
  tibble::tibble(
    model_id         = model_id,
    Season_AICc      = MuMIn::AICc(model),
    Temperature_AICc = if (is.null(m_temp)) NA_real_ else MuMIn::AICc(m_temp))
}) %>%
  dplyr::mutate(delta_AICc = Temperature_AICc - Season_AICc)

knitr::kable(season_vs_temperature, digits = 1)

```

## RQ1 alternative models check
```{r}
#| label: tbl-rq1-covariate-selection
#| tbl-cap: "AICc covariate selection for RQ1 baseline models (round 0, no random effects). Base: response ~ Site_type_detail. Season cannot vary within a single round. ΔAICc relative to best per model; supported = ΔAICc < 2."

# Shared helper — fits base + each named candidate with REML=FALSE and returns AICc table.
# model_id is NOT included in the output — the caller's parent tibble already has it.
covariate_select <- function(base_form, candidates, data, family, zif = ~0, focal_term = NULL) {
  fit_one <- function(form) {
    m <- tryCatch(
      glmmTMB::glmmTMB(as.formula(form), data = data, REML = FALSE,
                       family = family, ziformula = zif),
      error = function(e) NULL)
    if (!is.null(m) && m$fit$convergence != 0)
      m <- tryCatch(
        glmmTMB::glmmTMB(as.formula(form), data = data, REML = FALSE,
                         family = family, ziformula = zif,
                         control = glmmTMBControl(optCtrl = list(iter.max = 1000, eval.max = 2000))),
        error = function(e) NULL)
    m
  }
  all_forms <- c(list(Base = base_form), candidates)
  models    <- purrr::map(all_forms, fit_one)
  purrr::imap_dfr(models, function(m, name) {
    hess_ok <- if (is.null(m) || m$fit$convergence != 0) FALSE else
      tryCatch({ v <- diag(as.matrix(vcov(m, full = TRUE))); all(is.finite(v) & v > 0) },
               error = function(e) FALSE, warning = function(w) FALSE)
    se_focal <- if (!is.null(m) && !is.null(focal_term))
      tryCatch(
        broom.mixed::tidy(m, effects = "fixed", component = "cond") %>%
          dplyr::filter(term == focal_term) %>%
          dplyr::pull(std.error) %>%
          { if (length(.) == 0) NA_real_ else .[[1]] },
        error = function(e) NA_real_)
      else NA_real_
    r2m <- if (!is.null(m))
      tryCatch(MuMIn::r.squaredGLMM(m)[1, "R2m"], error = function(e) NA_real_)
      else NA_real_
    tibble::tibble(
      candidate  = name,
      K          = if (!is.null(m)) attr(logLik(m), "df") else NA_integer_,
      AICc       = if (!is.null(m)) MuMIn::AICc(m)       else NA_real_,
      SE_focal   = se_focal,
      R2m        = r2m,
      hessian_ok = hess_ok)
  }) %>%
    dplyr::mutate(delta_AICc = AICc - min(AICc, na.rm = TRUE),
                  supported  = delta_AICc < 2) %>%
    dplyr::arrange(delta_AICc)
}

# RQ1a — Stone_pile vs Control, round 0
rq1a_cov <- tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
  dplyr::mutate(
    model_id = paste0("M1a", response_letter_map[response_type], species_letter_map[species]),
    table    = purrr::pmap(list(species, response_type), function(sp, rt) {
      rc  <- if (rt == "abundance") paste0("Weighted_CPUE_", sp) else paste0("Weighted_BPUE_", sp)
      if (!rc %in% names(rq1a_data)) return(NULL)
      zif  <- if (sp == "Catfish" && rt == "abundance") ~1 else ~0
      base <- paste0("`", rc, "` ~ Site_type_detail")
      covariate_select(base,
        list(Temperature = paste0(base, " + Temperature"),
             Substrate   = paste0(base, " + Substrate_index_before"),
             Fetch       = paste0(base, " + fetch_z")),
        rq1a_data, tweedie(), zif,
        focal_term = "Site_type_detailStone_pile")
    }))

# RQ1b — Ref1 vs Ref2, round 0
rq1b_cov <- tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
  dplyr::mutate(
    model_id = paste0("M1b", response_letter_map[response_type], species_letter_map[species]),
    table    = purrr::pmap(list(species, response_type), function(sp, rt) {
      rc  <- if (rt == "abundance") paste0("Weighted_CPUE_", sp) else paste0("Weighted_BPUE_", sp)
      if (!rc %in% names(rq1b_data)) return(NULL)
      zif  <- if (sp == "Catfish" && rt == "abundance") ~1 else ~0
      base <- paste0("`", rc, "` ~ Site_type_detail")
      covariate_select(base,
        list(Temperature = paste0(base, " + Temperature"),
             Substrate   = paste0(base, " + Substrate_index_before"),
             Fetch       = paste0(base, " + fetch_z")),
        rq1b_data, tweedie(), zif,
        focal_term = "Site_type_detailReference_2")
    }))

rq1_cov_summary <- dplyr::bind_rows(
  tidyr::unnest(dplyr::filter(rq1a_cov, !purrr::map_lgl(table, is.null)), table),
  tidyr::unnest(dplyr::filter(rq1b_cov, !purrr::map_lgl(table, is.null)), table)) %>%
  dplyr::select(model_id, species, response_type, candidate, K, AICc, delta_AICc, SE_focal, R2m, supported, hessian_ok) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 2)))

print(rq1_cov_summary, n = 32)

DT::datatable(rq1_cov_summary, filter = "top",
  options = list(pageLength = 30, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons', rownames = FALSE)
```


## RQ2 alternative models check
```{r}
#| label: tbl-rq2-covariate-selection
#| tbl-cap: "AICc covariate selection for RQ2 models. RQ2a base: response ~ Period * site + monitoring_int + (1|pair_id/site_id). RQ2b base: response ~ site * monitoring_int + (1|site_id). One covariate added at a time."

# RQ2a — BACI step-change (full timeline, paired random effects)
rq2a_cov <- tidyr::crossing(species = all_species, response_type = c("occupancy", "abundance", "biomass")) %>%
  dplyr::mutate(
    model_id = paste0("M2a", response_letter_map[response_type], species_letter_map[species]),
    table    = purrr::pmap(list(species, response_type), function(sp, rt) {
      rc <- dplyr::case_when(
        rt == "occupancy" ~ paste0("Presence_", sp),
        rt == "abundance" ~ paste0("Weighted_CPUE_", sp),
        rt == "biomass"   ~ paste0("Weighted_BPUE_", sp))
      if (!rc %in% names(rq2a_data)) return(NULL)
      fam  <- if (rt == "occupancy") binomial() else tweedie()
      zif  <- if (sp == "Catfish" && rt == "abundance") ~1 else ~0
      base <- paste0("`", rc, "` ~ Period * site + monitoring_int + (1 | pair_id / site_id)")
      covariate_select(base,
        list(Temperature = paste0(base, " + Temperature"),
             Season      = paste0(base, " + Season"),
             Substrate   = paste0(base, " + Substrate_index_before"),
             Fetch       = paste0(base, " + fetch_z"),
             Wave        = paste0(base, " + wave_z")),
        rq2a_data, fam, zif,
        focal_term = "PeriodAfter:siteStone_pile")
    }))

# RQ2b — post-construction trajectory (rounds 1-6, site_id random effects)
rq2b_cov <- tidyr::crossing(species = all_species, response_type = c("occupancy", "abundance", "biomass")) %>%
  dplyr::mutate(
    model_id = paste0("M2b", response_letter_map[response_type], species_letter_map[species]),
    table    = purrr::pmap(list(species, response_type), function(sp, rt) {
      rc <- dplyr::case_when(
        rt == "occupancy" ~ paste0("Presence_", sp),
        rt == "abundance" ~ paste0("Weighted_CPUE_", sp),
        rt == "biomass"   ~ paste0("Weighted_BPUE_", sp))
      if (!rc %in% names(rq2b_data)) return(NULL)
      fam  <- if (rt == "occupancy") binomial() else tweedie()
      zif  <- if (sp == "Catfish" && rt == "abundance") ~1 else ~0
      base <- paste0("`", rc, "` ~ site * monitoring_int + (1 | site_id)")
      covariate_select(base,
        list(Temperature = paste0(base, " + Temperature"),
             Season      = paste0(base, " + Season"),
             Substrate   = paste0(base, " + Substrate_index_before"),
             Fetch       = paste0(base, " + fetch_z"),
             Wave        = paste0(base, " + wave_z")),
        rq2b_data, fam, zif,
        focal_term = "siteStone_pile:monitoring_int")
    }))

rq2_cov_summary <- dplyr::bind_rows(
  tidyr::unnest(dplyr::filter(rq2a_cov, !purrr::map_lgl(table, is.null)), table),
  tidyr::unnest(dplyr::filter(rq2b_cov, !purrr::map_lgl(table, is.null)), table)) %>%
  dplyr::select(model_id, species, response_type, candidate, K, AICc, delta_AICc, SE_focal, R2m, supported, hessian_ok) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 2)))

print(rq2_cov_summary, n = 48)

DT::datatable(rq2_cov_summary, filter = "top",
  options = list(pageLength = 40, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons', rownames = FALSE)
```


## RQ3 alternative models check
```{r}
#| label: tbl-rq3-covariate-selection
#| tbl-cap: "AICc covariate selection for RQ3 models. RQ3a base: response ~ site * monitoring_int + (1|site_id). RQ3b base: response ~ Site_type_detail * monitoring_int + (1|pair_id/site_id). One covariate added at a time."

# RQ3a — convergence (Stone_pile vs Reference, rounds 1-6)
rq3a_cov <- tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
  dplyr::mutate(
    model_id = paste0("M3a", response_letter_map[response_type], species_letter_map[species]),
    table    = purrr::pmap(list(species, response_type), function(sp, rt) {
      rc  <- if (rt == "abundance") paste0("Weighted_CPUE_", sp) else paste0("Weighted_BPUE_", sp)
      if (!rc %in% names(rq3a_data)) return(NULL)
      zif  <- if (sp == "Catfish" && rt == "abundance") ~1 else ~0
      base <- paste0("`", rc, "` ~ site * monitoring_int + (1 | site_id)")
      covariate_select(base,
        list(Temperature     = paste0(base, " + Temperature"),
             Season          = paste0(base, " + Season"),
             Substrate_after = paste0(base, " + Substrate_index_after"),
             Fetch           = paste0(base, " + fetch_z"),
             Wave            = paste0(base, " + wave_z")),
        rq3a_data, tweedie(), zif,
        focal_term = "siteStone_pile:monitoring_int")
    }))

# RQ3b — reference stability (Ref1 vs Ref2, rounds 1-6)
rq3b_cov <- tidyr::crossing(species = all_species, response_type = c("abundance", "biomass")) %>%
  dplyr::mutate(
    model_id = paste0("M3b", response_letter_map[response_type], species_letter_map[species]),
    table    = purrr::pmap(list(species, response_type), function(sp, rt) {
      rc  <- if (rt == "abundance") paste0("Weighted_CPUE_", sp) else paste0("Weighted_BPUE_", sp)
      if (!rc %in% names(rq3b_data)) return(NULL)
      zif  <- if (sp == "Catfish" && rt == "abundance") ~1 else ~0
      base <- paste0("`", rc, "` ~ Site_type_detail * monitoring_int + (1 | pair_id / site_id)")
      covariate_select(base,
        list(Temperature = paste0(base, " + Temperature"),
             Season      = paste0(base, " + Season"),
             Fetch       = paste0(base, " + fetch_z"),
             Wave        = paste0(base, " + wave_z")),
        rq3b_data, tweedie(), zif,
        focal_term = "Site_type_detailReference_2:monitoring_int")
    }))

rq3_cov_summary <- dplyr::bind_rows(
  tidyr::unnest(dplyr::filter(rq3a_cov, !purrr::map_lgl(table, is.null)), table),
  tidyr::unnest(dplyr::filter(rq3b_cov, !purrr::map_lgl(table, is.null)), table)) %>%
  dplyr::select(model_id, species, response_type, candidate, K, AICc, delta_AICc, SE_focal, R2m, supported, hessian_ok) %>%
  dplyr::mutate(across(where(is.numeric), ~ round(.x, 2)))

print(rq3_cov_summary, n = 28)

DT::datatable(rq3_cov_summary, filter = "top",
  options = list(pageLength = 30, dom = 'Bfrtip', buttons = c('csv', 'excel')),
  extensions = 'Buttons', rownames = FALSE)
```

## Habitat compression model
```{r}
#| label: tbl-habitat-compression
# Does kōura abundance at stone pile sites increase relative to control/reference
# sites as habitable depth decreases (thermal compression hypothesis)?

hc_data <- Combined_data |>
  dplyr::filter(
    monitoring > 0,
    is.finite(habitable_depth_m),
    !is.na(`Weighted_CPUE_Kōura`)) |>
  dplyr::mutate(
    site = factor(site, levels = c("Control", "Stone_pile", "Reference")),
    habitable_depth_c = habitable_depth_m - mean(habitable_depth_m, na.rm = TRUE)  )

m_hc_koura <- glmmTMB::glmmTMB(
  `Weighted_CPUE_Kōura` ~ site * habitable_depth_c + monitoring_int + (1 | site_id),
  data   = hc_data,
  family = glmmTMB::tweedie(link = "log"))

summary(m_hc_koura)

# Tidy results for inline use
hc_tidy <- broom.mixed::tidy(m_hc_koura, conf.int = TRUE, exponentiate = FALSE) |>
  dplyr::filter(effect == "fixed") |>
  dplyr::select(term, estimate, std.error, statistic, p.value, conf.low, conf.high)

hc_tidy
```

## Habitat predictor variables
```{r}
# Habitat predictor model — all sites, post-construction
hab_model_data <- all_data |>
  dplyr::filter(monitoring_int >= 1, !is.na(`Weighted_CPUE_Kōura`)) |>
  dplyr::mutate(
    site_id = factor(site_id),
    # centre continuous predictors
    wood_c       = scale(Wood_cover)[,1],
    trees_c      = scale(Overhanging_trees)[,1],
    riparian_c   = scale(Riparian_vegetation)[,1],
    substrate_c  = scale(Substrate_index_after)[,1],
    slope5_c     = scale(Slope_0_5m)[,1],
    slope20_c    = scale(Slope_0_20m)[,1])

# Fit each habitat variable individually (AICc comparison)
library(glmmTMB)

hab_vars <- c("wood_c", "trees_c", "riparian_c", "substrate_c", "slope5_c", "slope20_c")

hab_null <- glmmTMB(`Weighted_CPUE_Kōura` ~ monitoring_int + (1 | site_id),
                    data = hab_model_data, REML = FALSE, family = tweedie())

hab_fits <- purrr::map(hab_vars, \(v) {
  form <- paste0("`Weighted_CPUE_Kōura` ~ ", v, " + monitoring_int + (1 | site_id)")
  glmmTMB(as.formula(form), data = hab_model_data, REML = FALSE, family = tweedie())
}) |> purrr::set_names(hab_vars)

# AICc table
purrr::map_dfr(hab_vars, \(v) {
  m <- hab_fits[[v]]
  tibble::tibble(
    variable   = v,
    AICc       = MuMIn::AICc(m),
    delta_AICc = MuMIn::AICc(m) - MuMIn::AICc(hab_null),
    beta       = broom.mixed::tidy(m, effects = "fixed") |>
                   dplyr::filter(stringr::str_detect(term, v)) |>
                   dplyr::pull(estimate),
    p          = broom.mixed::tidy(m, effects = "fixed") |>
                   dplyr::filter(stringr::str_detect(term, v)) |>
                   dplyr::pull(p.value))
}) |> dplyr::arrange(AICc)



# One row per site — clean site-level summary
site_summary <- all_data |>
  dplyr::filter(monitoring_int >= 1, !is.na(`Weighted_CPUE_Kōura`)) |>
  dplyr::group_by(site_id, site, pair_id) |>
  dplyr::summarise(
    mean_cpue          = mean(`Weighted_CPUE_Kōura`, na.rm = TRUE),
    substrate_after    = max(Substrate_index_after),
    wood_cover         = mean(Wood_cover, na.rm = TRUE),
    overhanging_trees  = mean(Overhanging_trees, na.rm = TRUE),
    riparian_veg       = mean(Riparian_vegetation, na.rm = TRUE),
    slope_20           = mean(Slope_0_20m, na.rm = TRUE),
    slope_5            = mean(Slope_0_5m, na.rm = TRUE),
    .groups = "drop")

# Correlate everything
site_summary |>
  dplyr::select(mean_cpue, substrate_after, wood_cover, 
                overhanging_trees, riparian_veg, slope_20, slope_5) |>
  cor(method = "spearman")

# Scale predictors
site_scaled <- site_summary |>
  dplyr::mutate(across(substrate_after:slope_5, scale))

# Full model
m_full <- lm(log1p(mean_cpue) ~ substrate_after + wood_cover + 
               overhanging_trees + riparian_veg + slope_20, 
             data = site_scaled,
             na.action = na.fail)

MuMIn::dredge(m_full)

```

# Interactive widgets (saved as RDS for HTML thesis)

```{r}
#| label: widget-forest-plot-ch5
#| include: false
widget_forest_rds <- file.path(der_data_dir, "widget_forest_plot.rds")
if (!file.exists(widget_forest_rds)) {
  grp_short <- c(
    "Were stone pile and control sites\ncomparable before construction?"      = "Baseline: SP vs Control",
    "Were the two reference groups\ncomparable at baseline?"                  = "Baseline: Ref groups",
    "Was the before-after change greater at\nstone pile than control sites?"  = "BACI: SP vs Control",
    "Did stone pile and control sites diverge\nover the post-construction period?" = "Trend: SP vs Control",
    "Did stone pile sites converge toward\nreference site densities over time?"    = "Convergence: SP vs Ref",
    "Were reference site conditions stable\nacross monitoring rounds?"              = "Reference stability"
  )
  fp_w <- forest_data |>
    dplyr::mutate(
      grp_s  = grp_short[as.character(group)],
      y_lbl  = paste0(grp_s, " | ", row_label),
      hover  = paste0(
        "<b>", species, " — ", row_label, "</b><br>",
        grp_s, "<br>",
        "β = ", round(estimate, 3),
        " [", round(conf.low, 3), ", ", round(conf.high, 3), "]<br>",
        "p = ", signif(p.value, 3), ifelse(significant, " ★", "")
      )
    )

  make_fp <- function(sp_data, show_y = TRUE) {
    plotly::plot_ly(sp_data) |>
      plotly::add_segments(
        x = ~conf.low, xend = ~conf.high, y = ~y_lbl, yend = ~y_lbl,
        line = list(color = "rgba(60,60,60,0.65)", width = 1.8),
        hoverinfo = "none", showlegend = FALSE
      ) |>
      plotly::add_markers(
        x = ~estimate, y = ~y_lbl,
        marker = list(
          size   = 9,
          color  = ~ifelse(significant, "#1a1a1a", "white"),
          symbol = "circle",
          line   = list(color = "#1a1a1a", width = 1.5)
        ),
        text = ~hover, hoverinfo = "text", showlegend = FALSE
      ) |>
      plotly::layout(
        shapes = list(list(
          type = "line", x0 = 0, x1 = 0, y0 = 0, y1 = 1,
          yref = "paper", line = list(color = "grey60", dash = "dash", width = 1)
        )),
        xaxis = list(title = "β ± 95% CI", zeroline = FALSE, showgrid = TRUE),
        yaxis = list(title = "", autorange = "reversed",
                     showticklabels = show_y, tickfont = list(size = 11))
      )
  }

  p_k <- make_fp(dplyr::filter(fp_w, species == "Kōura"),   show_y = TRUE)
  p_c <- make_fp(dplyr::filter(fp_w, species == "Catfish"), show_y = FALSE)

  fp_widget <- plotly::subplot(p_k, p_c, nrows = 1, shareY = TRUE, titleX = TRUE,
                               widths = c(0.6, 0.4)) |>
    plotly::layout(
      annotations = list(
        list(x = 0.28, y = 1.04, text = "<b>Kōura</b>",   showarrow = FALSE,
             xref = "paper", yref = "paper", font = list(size = 13)),
        list(x = 0.82, y = 1.04, text = "<b>Catfish</b>", showarrow = FALSE,
             xref = "paper", yref = "paper", font = list(size = 13))
      ),
      margin = list(t = 60)
    )
  saveRDS(fp_widget, widget_forest_rds)
}
```


```{r}
#| label: widget-site-overview-ch5
#| include: false
widget_site_ov_rds <- file.path(der_data_dir, "widget_site_overview.rds")
if (!file.exists(widget_site_ov_rds)) {
  site_ov_w <- readr::read_csv(file.path(out_dir, "tbl-site-overview.csv"), show_col_types = FALSE)
  dt_w <- DT::datatable(site_ov_w, rownames = FALSE, filter = "top",
                        options = list(pageLength = 25, dom = 'frtip', scrollX = TRUE)) |>
    DT::formatRound(columns = which(sapply(site_ov_w, is.numeric)), digits = 2)
  saveRDS(dt_w, widget_site_ov_rds)
}
```

```{r}
#| label: widget-sites-map-ch5
#| include: false
widget_sites_map_rds <- file.path(der_data_dir, "widget_sites_map.rds")
if (!file.exists(widget_sites_map_rds)) {
  rotoiti_w <- sf::st_read("data/raw/Lake Rotoiti.gpkg", quiet = TRUE) |>
    sf::st_make_valid() |>
    sf::st_transform(4326)
  site_info_w <- readr::read_csv(file.path(der_data_dir, "Site_info.csv"), show_col_types = FALSE) |>
    dplyr::filter(monitoring == 0) |>
    dplyr::distinct(site_id, site, lat, lon)
  site_colors_w <- c("Stone_pile" = "#c0392b", "Control" = "#e67e22", "Reference" = "#2980b9")
  site_info_w <- dplyr::mutate(site_info_w,
    color = dplyr::case_when(
      grepl("stone",     site, ignore.case = TRUE) ~ "#c0392b",
      grepl("control",   site, ignore.case = TRUE) ~ "#e67e22",
      grepl("reference", site, ignore.case = TRUE) ~ "#2980b9",
      TRUE ~ "#555555"
    ),
    site_label = dplyr::recode(site, "Stone_pile" = "Stone pile")
  )
  sites_map_w <- leaflet::leaflet() |>
    leaflet::addProviderTiles("Esri.WorldGrayCanvas") |>
    leaflet::addPolygons(
      data = rotoiti_w, fillColor = "#bbdefb", fillOpacity = 0.45,
      color = "#1565c0", weight = 1.5, label = "Lake Rotoiti"
    ) |>
    leaflet::addCircleMarkers(
      data = site_info_w, lat = ~lat, lng = ~lon, radius = 7,
      color = ~color, fillColor = ~color, fillOpacity = 0.9, weight = 1.5,
      popup = ~paste0("<b>Site ", site_id, "</b><br>Type: ", site_label)
    ) |>
    leaflet::addLegend(position = "bottomright", colors = unname(site_colors_w),
                       labels = c("Stone pile", "Control", "Reference"),
                       title = "Site type", opacity = 0.9)
  saveRDS(sites_map_w, widget_sites_map_rds)
}
```

# Session info
```{r}
#| label: session-info
#| include: true
#| code-fold: true
sessionInfo()
```